---
title: "Chapter 3: *Some Basic Probability Concepts*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp03}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette of package **`DanielBiostatistics10th`** ([CRAN](https://cran.r-project.org/package=DanielBiostatistics10th), [Github](https://github.com/tingtingzhan/DanielBiostatistics10th)) documents the examples of Chapter 3 *Some Basic Probability Concepts* of 
***Biostatistics: A Foundation for Analysis in the Health Sciences***, 
[11th Edition](https://www.wiley.com/en-us/Biostatistics%3A+A+Foundation+for+Analysis+in+the+Health+Sciences%2C+11th+Edition-p-9781119496571) by Wayne W. Daniel and Chad L. Cross.

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(DanielBiostatistics10th)
```

## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', 'https://cran.r-project.org',
  '`flextable`', 'Flexible tables', '`?flextable::flextable`',
  '`matrix`', 'Matrices', '`?base::matrix`',
  '`table`', 'Cross tabulation', '`?base::table`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable(format = 'html') 
# ?knitr::kable
# default: `|` shown as &...
# format = 'html': `>` shown as &..
```




# Example 3.4.1 - 3.4.9
Page 69-75 (10th ed), Page 61-67 (11th ed)
```{r}
c(28L, 19L, 41L, 53L, 35L, 38L, 44L, 60L) |>
  matrix(ncol = 2L, dimnames = list(
    FamilyHx = c('none', 'Bipolar', 'Unipolar', 'UniBipolar'), 
    Onset = c('Early', 'Late')
  )) |> 
  as.table() |>
  as_flextable()
```

# Example 3.5.1
Page 81 (10th ed), Page 72 (11th ed)
```{r}
d351 = c(495L, 14L, 5L, 436L) |>
  matrix(nrow = 2L, dimnames = list(
    Alzheimer = c('No', 'Yes'), 
    Test = c('Negative', 'Positive')
  )) |>
  binTab()
d351 |>
  as_flextable()
```
```{r}
d351 |>
  print(prevalence = .113, print_flextable = FALSE)
```