---
title: "Chapter 1: *Introduction to Biostatistics*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp01}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette of package **`DanielBiostatistics10th`** ([CRAN](https://cran.r-project.org/package=DanielBiostatistics10th), [Github](https://github.com/tingtingzhan/DanielBiostatistics10th)) documents the examples of Chapter 1 *Introduction to Biostatistics* of 
***Biostatistics: A Foundation for Analysis in the Health Sciences***, 
[11th Edition](https://www.wiley.com/en-us/Biostatistics%3A+A+Foundation+for+Analysis+in+the+Health+Sciences%2C+11th+Edition-p-9781119496571) by Wayne W. Daniel and Chad L. Cross.

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(DanielBiostatistics10th)
```

## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`$`', 'Extract parts of an object', '`` ?base::`$` ``',
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', 'https://cran.r-project.org',
  '`class`', 'Object class', '`?base::class`',
  '`data.frame`', 'Data frame', '`?base::data.frame`',
  '`dim`', 'Dimensions', '`?base::dim`',
  '`head`, `tail`', 'First or last parts of an object', '`?utils::head`',
  '`names`', 'Names of an object', '`?base::names`',
  '`nrow`', 'Number of rows', '`?base::nrow`',
  '`sample`, `sample.int`', 'Random samples', '`?base::sample`; `?base::sample.int`', 
  '`search`', 'Search path', '`?base::search`',
  '`seed`', 'Random number generation seed', '`?base::set.seed`',
  '`seq`, `seq.int`', 'Sequence generation', '`?base::seq`; `?base::seq.int`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable(format = 'html') 
# ?knitr::kable
# default: `|` shown as &...
# format = 'html': `>` shown as &..
```


# Example 1.4.1

Page 8 (10th ed), Page 7 (11th ed).
```{r}
EXA_C01_S04_01 |> 
  class() 
```
```{r}
EXA_C01_S04_01 |>
  dim()
```
```{r}
EXA_C01_S04_01 |> 
  head(n = 3L)
```
```{r}
EXA_C01_S04_01 |> 
  names()
```
```{r}
EXA_C01_S04_01$AGE
```

To answer Example 1.4.1,
```{r}
(nr = EXA_C01_S04_01 |> nrow())
set.seed(142); EXA_C01_S04_01[sample.int(n = nr, size = 10L), ]
```

# Example 1.4.2

Page 11 (10th ed), Page 10 (11th ed).
```{r}
EXA_C01_S04_01[seq.int(from = 4L, to = 166L, by = 18L), ]
```