---
title: "Chapter 6, *Estimation*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp06}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(DanielBiostatistics10th)
```




# Example 6.2.1
Page 166 (10th ed), Page 147 (11th ed)
```{r}
aggregated_z(xbar = 22, n = 10L, sd = sqrt(45))
```

# Example 6.2.2
Page 168 (10th ed), Page 149 (11th ed)
```{r}
aggregated_z(xbar = 84.3, n = 15L, sd = sqrt(144), conf.level = .99)
```

# Example 6.2.3
Page 168 (10th ed), Page 150 (11th ed)
```{r}
aggregated_z(xbar = 17.2, n = 35L, sd = 8, conf.level = .9)
```

# Example 6.2.4
Page 169 (10th ed), Page 150 (11th ed)
```{r}
head(EXA_C06_S02_04)
aggregated_z(xbar = mean(EXA_C06_S02_04$ACTIVITY), n = nrow(EXA_C06_S02_04), sd = sqrt(.36))
```

# Example 6.3.1
Page 173 (10th ed), 
```{r}
aggregated_t(xbar = 250.8, xsd = 130.9, n = 19L)
```

# Example 6.4.1
Page 177 (10th ed), 
```{r}
aggregated_z(xbar = c(4.5, 3.4), sd = sqrt(c(1, 1.5)), n = c(12L, 15L))
```

# Example 6.4.2
Page 178 (10th ed), 
```{r}
aggregated_z(xbar = c(4.3, 13), sd = c(5.22, 8.97), n = c(328L, 64L), conf.level = .99)
```

# Example 6.4.3
Page 180 (10th ed), 
```{r}
aggregated_t(xbar = c(4.7, 8.8), xsd = c(9.3, 11.5), n = c(18L, 10L), var.equal = TRUE)
```

# Example 6.4.4
Page 181 (10th ed), 
```{r}
aggregated_t(xbar = c(4.7, 8.8), xsd = c(9.3, 11.5), n = c(18L, 10L))
```

Welch slightly different from Cochran; textbook explained on Page 182

# Example 6.5.1
Page 185 (10th ed), 
```{r}
prop_CLT(xbar = .18, n = 1220L)
```

# Example 6.6.1
Page 187 (10th ed), 
```{r}
prop_CLT(x = c(31L, 53L), n = c(68L, 255L), conf.level = .99)
```

# Example 6.7.1
Page 190 (10th ed), 
```{r}
uniroot(f = function(n, sd, level = .95) {
  qnorm(1-(1-level)/2) * sd/sqrt(n) - 5 # half-width of CI <= 5 grams
}, interval = c(0, 2e2), sd = 20)$root |>
  ceiling()
```

# Example 6.8.1
Page 192 (10th ed), 
```{r}
uniroot(f = function(n, p, level = .95) {
  qnorm(1-(1-level)/2) * sqrt(p*(1-p)/n) - .05
}, interval = c(0, 1e3), p = .35)$root |>
  ceiling()
```

# Example 6.9.1
Page 196 (10th ed), 
```{r}
d691 = c(9.7, 12.3, 11.2, 5.1, 24.8, 14.8, 17.7)
sqrt(aggregated_var(xsd = sd(d691), n = length(d691)))
```

# Example 6.10.1
Page 200 (10th ed), 
```{r}
aggregated_var(xsd = c(8.1, 5.9), n = c(16L, 4L))
```
