---
title: "Chapter 2: *Descriptive Statistics*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp02}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette of package **`DanielBiostatistics10th`** ([CRAN](https://cran.r-project.org/package=DanielBiostatistics10th), [Github](https://github.com/tingtingzhan/DanielBiostatistics10th)) documents the examples of Chapter 2 *Descriptive Statistics* of 
***Biostatistics: A Foundation for Analysis in the Health Sciences***, 
[11th Edition](https://www.wiley.com/en-us/Biostatistics%3A+A+Foundation+for+Analysis+in+the+Health+Sciences%2C+11th+Edition-p-9781119496571) by Wayne W. Daniel and Chad L. Cross.

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(DanielBiostatistics10th)
```

## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', 'https://cran.r-project.org',
  '`cut`', 'Convert numeric to factor', '`?base::cut`; `?base::cut.default`',
  '`factor`', 'Factor', '`?base::factor`',
  '`kurtosis`', 'Kurtosis', '`?e1071::kurtosis`',
  '`mean`', 'Arithmetic mean', '`?base::mean`; `?base::mean.default`',
  '`median`', 'Median', '`?stats::median`; `?stats::median.default`',
  '`Mode`', 'Mode', '`?pracma::Mode`',
  '`quantile`', 'Quantile', '`?stats::quantile`',
  '`sd`', 'Standard deviation', '`?stats::sd`',
  '`skewness`', 'Skewness', '`?e1071::skewness`',
  '`sort`', 'Sorting vectors', '`?base::sort`',
  '`var`', 'Variance', '`?stats::var`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable(format = 'html') 
# ?knitr::kable
# default: `|` shown as &...
# format = 'html': `>` shown as &..
```




# Example 2.2.1
Page 20 (10th ed), Page 19 (11th ed)
```{r}
EXA_C01_S04_01 |> 
  head(n = 3L)
```
```{r}
EXA_C01_S04_01$AGE |> 
  class()
```

Table 2.2.1.
```{r}
EXA_C01_S04_01$AGE |> 
  sort()
```


# Example 2.3.1

Page 23 (10th ed); Page 22 (11th ed).

Table 2.3.2
```{r}
EXA_C01_S04_01$AGE |>
  cut(breaks = seq.int(from=30, to=90, by=10), right = FALSE) |>
  viewFreq()
```




# Example 2.4.1-2.4.6; 2.5.1-2.5.3

Page 38-42 (10th ed); Page 30-34 (11th ed).

Page 44-46 (10th ed); Page 35-41 (11th ed).
```{r}
EXA_C01_S04_01$AGE |> 
  viewStats()
```

# Example 2.5.4 (omitted)
Page 49 (10th ed), Page 41 (11th ed)

# Example 2.5.5
Page 50 (10th ed)
```{r}
EXA_C02_S05_05 |>
  head(n = 3L)
```
```{r}
EXA_C02_S05_05$GRF |>
  boxplot(main = c('Example 2.5.5 (10th ed)'))
```
```{r}
EXA_C02_S05_05$GRF |> 
  viewStats()
```
