---
title: "Chapter 4: *Probability Distributions*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp04}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```


# Introduction

This vignette of package **`DanielBiostatistics10th`** ([CRAN](https://cran.r-project.org/package=DanielBiostatistics10th), [Github](https://github.com/tingtingzhan/DanielBiostatistics10th)) documents the examples of Chapter 2 *Descriptive Statistics* of 
***Biostatistics: A Foundation for Analysis in the Health Sciences***, 
[11th Edition](https://www.wiley.com/en-us/Biostatistics%3A+A+Foundation+for+Analysis+in+the+Health+Sciences%2C+11th+Edition-p-9781119496571) by Wayne W. Daniel and Chad L. Cross.

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(DanielBiostatistics10th)
```

## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', 'https://cran.r-project.org',
  '`dbinom`, `pbinom`', 'Binomial density and distribution', '`?stats::dbinom`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable(format = 'html') 
# ?knitr::kable
# default: `|` shown as &...
# format = 'html': `>` shown as &..
```




# Example 4.2.1-4.2.7

Page 93-97 (10th ed), Page 81-85 (11th ed)
```{r}
(d421 = rep(1:8, times = c(62L, 47L, 39L, 39L, 58L, 37L, 4L, 11L)))
```

Table 4.2.1, 4.2.2, Table 4.2.3
```{r}
d421 |>
  ordered() |>
  viewFreq()
```

# Example 4.2.8
Page 98 (10th ed), Page 85 (11th ed).
```{r}
viewStats(d421)
```



# Example 4.3.1

Page 99 (10th ed).

`d` for `density`; calculates $\text{Pr}(X = x)$.

```{r}
dbinom(x = 3L, size = 5L, prob = .858)
```

# Example 4.3.1
Page 87 (11th ed).
```{r}
dbinom(x = 3L, size = 5L, prob = .899)
```

# Example 4.3.2
Page 103 (10th ed), Page 90 (11th ed).
```{r}
dbinom(x = 4L, size = 10L, prob = .14)
```

# Example 4.3.3
Page 103 (10th ed), Page 91 (11th ed).

`p` for `probability` or `distribution`; calculates

-   $\text{Pr}(X \leq x)$ if `lower.tail = TRUE` (default)
-   $\text{Pr}(X > x)$ if `lower.tail = FALSE`

```{r}
(pL = pbinom(q = 5L, size = 25L, prob = .1, lower.tail = TRUE))
(pU = pbinom(q = 5L, size = 25L, prob = .1, lower.tail = FALSE))
stopifnot(pL + pU == 1)
```

# Example 4.3.4
Page 105 (10th ed), Page 92 (11th ed) 
```{r}
dbinom(x = 7L, size = 12L, prob = .55)
pbinom(q = 5L, size = 12L, prob = .55)
pbinom(q = 7L, size = 12L, prob = .55, lower.tail = FALSE)
```

# Example 4.4.1
Page 110 (10th ed), Page 97 (11th ed) 
```{r}
dpois(x = 3L, lambda = 12) 
```

# Example 4.4.2
Page 110 (10th ed), Page 98 (11th ed) 
```{r}
ppois(q = 2L, lambda = 12, lower.tail = FALSE)
```

# Example 4.4.3
Page 110 (10th ed), Page 98 (11th ed) 
```{r}
ppois(q = 1L, lambda = 2) 
```

# Example 4.4.4
Page 111 (10th ed), Page 98 (11th ed) 
```{r}
dpois(x = 3L, lambda = 2)
```

# Example 4.4.5
Page 112 (10th ed), Page 98 (11th ed) 
```{r}
ppois(q = 5L, lambda = 2, lower.tail = FALSE)
```

# Example 4.6.1
Page 119 (10th ed), Page 106 (11th ed) 
```{r}
pnorm(q = 2)
```

# Example 4.6.2
Page 120 (10th ed), Page 106 (11th ed) 
```{r}
pnorm(2.55) - pnorm(-2.55)
1 - 2 * pnorm(-2.55) # alternative solution
```

# Example 4.6.3
Page 121 (10th ed), Page 107 (11th ed) 
```{r}
pnorm(1.53) - pnorm(-2.74)
```

# Example 4.6.4
Page 121 (10th ed), Page 107 (11th ed) 
```{r}
pnorm(2.71, lower.tail = FALSE)
```

# Example 4.6.5
Page 122 (10th ed), Page 107 (11th ed) 
```{r}
pnorm(2.45) - pnorm(.84)
```

# Example 4.7.1
Page 122 (10th ed), Page 109 (11th ed) 
```{r}
pnorm(q = 3, mean = 5.4, sd = 1.3)
pnorm(q = (3-5.4)/1.3) # manual solution
```

# Example 4.7.2
Page 125 (10th ed), Page 111 (11th ed) 
```{r}
pnorm(q = 649, mean = 491, sd = 119) - pnorm(q = 292, mean = 491, sd = 119)
```

# Example 4.7.3
Page 122 (10th ed), Page 111 (11th ed) 
```{r}
1e4L * pnorm(q = 8.5, mean = 5.4, sd = 1.3, lower.tail = FALSE)
```