---
title: "Chapter 7: *Hypothesis Testing*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp07}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(DanielBiostatistics10th)
```

# Example 7.2.1
Page 222 (10th ed); Page 201 (11th ed)
```{r}
aggregated_z(xbar = 27, sd = sqrt(20), n = 10L, null.value = 30)
```

# Example 7.2.2
Page 226 (10th ed); Page 204 (11th ed)
```{r}
aggregated_z(xbar = 27, sd = sqrt(20), n = 10L, null.value = 30, alternative = 'less')
```

# Example 7.2.3
Page 228 (10th ed); Page 206 (11th ed)
```{r}
EXA_C07_S02_03 |> head(n = 3L)
t.test(EXA_C07_S02_03$DAYS, mu = 15)
```

# Example 7.2.4
Page 231 (10th ed); Page 209 (11th ed)
```{r}
aggregated_z(xbar = 146, sd = 27, n = 157L, null.value = 140, alternative = 'greater')
```

# Example 7.2.5
Page 232 (10th ed); Page 210 (11th ed)
```{r}
d725 = c(33.38, 32.15, 34.34, 33.95, 33.46, 34.13, 33.99, 34.10, 33.85, 
  34.23, 34.45, 34.19, 33.97, 32.73, 34.05)
t.test(d725, mu = 34.5)
```

# Example 7.3.1
Page 237 (10th ed), Page 213 (11th ed) 
```{r}
aggregated_z(xbar = c(4.5, 3.4), sd = sqrt(c(1, 1.5)), n = c(12L, 15L), null.value = 0)
```

# Example 7.3.2
Page 239 (10th ed), Page 215 (11th ed) 
```{r}
EXA_C07_S03_02 |> 
  head(n = 3L)
```
```{r}
EXA_C07_S03_02 |>
  with(t.test(x = CONTROL, y = SCI, alternative = 'less', var.equal = TRUE))
```

# Example 7.3.3
Page 240 (10th ed), Page 217 (11th ed) 
```{r}
aggregated_t(xbar = c(19.16, 9.53), xsd = c(5.29, 2.69), n = c(15L, 30L), null.value = 0)
```

# Example 7.3.4
Page 242 (10th ed), Page 219 (11th ed) 
```{r}
aggregated_z(xbar = c(59.01, 46.61), sd = c(44.89, 34.85), n = c(53L, 54L), null.value = 0,
  alternative = 'greater')
```

# Example 7.4.1
Page 251 (10th ed), Page 226 (11th ed) 
```{r}
EXA_C07_S04_01 |> head(n = 3L)
```
```{r}
EXA_C07_S04_01 |>
  with(t.test(x = POSTOP, y = PREOP, alternative = 'greater', paired = TRUE))
```

# Example 7.5.1
Page 258 (10th ed), Page 232 (11th ed) 
```{r}
prop_CLT(x = 24L, n = 301L, null.value = .063, alternative = 'greater') 
```

# Example 7.6.1
Page 261 (10th ed), Page 235 (11th ed) 
```{r}
prop_CLT(x = c(24L, 11L), n = c(44L, 29L), null.value = 0, alternative = 'greater')
```

# Example 7.7.1
Page 264 (10th ed), Page 238 (11th ed) 
```{r}
EXA_C07_S07_01 |> 
  head(n = 3L)
```
```{r}
aggregated_var(xsd = sd(EXA_C07_S07_01$mass), n = 16L, null.value = 600)
```

# Example 7.8.1
Page 268 (10th ed), Page 242 (11th ed) 
```{r}
aggregated_var(xsd = c(30.62, 11.37), n = 6L, null.value = 1, alternative = 'greater')
```

# Example 7.8.2
Page 270 (10th ed), 
```{r}
EXA_C07_S03_02 |>
  with(var.test(x = CONTROL, y = SCI))
```

# Example 7.9.1
Page 272 (10th ed), Page 245 (11th ed).
Table 7.9.1
```{r}
seq.int(from = 16, to = 19, by = .5) |>
  power_z(null.value = 17.5, sd = 3.6, n = 100L)
```

# Example 7.9.2
Page 276 (10th ed), Page 248 (11th ed) 
```{r}
seq.int(from = 50, to = 70, by = 5) |>
   power_z(null.value = 65, sd = 15, n = 20L, sig.level = .01, alternative = 'less')
```

# Example 7.10.1
Page 278 (10th ed), 
```{r}
(n_d7101 <- uniroot(f = function(x) {
  power_z(55, null.value = 65, sd = 15, n = x, sig.level = .01, alternative = 'less')$power - .95
}, interval = c(0, 50))$root)
power_z(55, null.value = 65, sd = 15, n = ceiling(n_d7101), sig.level = .01, alternative = 'less')
```