---
title: "Chapter 14: *Survival Analysis*"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chp14}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```


```{r setup}
library(DanielBiostatistics10th)
library(survival)
library(ggplot2)
library(ggfortify)
library(survminer)
```


### Example 14.3.1

Page 756 (10th ed), Page 652 (11th ed).

```{r}
EXA_C14_S03_01 |> head(n = 3L)
d1431 = EXA_C14_S03_01 |> 
  within(expr = {
    TIME = .difftime(TIME, units = 'months')
    OS = Surv(time = TIME, event = (VITAL != 'ned'))
  })
class(d1431$OS)
head(d1431$OS)
```

Table 14.3.2

```{r}
summary(sf_1431 <- survfit(OS ~ TUMOR, data = d1431))
```

Figure 14.3.1

```{r}
sf_1431 |> 
  plot(lty = 2:3, xlab = 'month', ylab = 'OS', main = 'Figure 14.3.1')
legend(x = 250, y = .9, legend = c('High', 'Low'), lty = 2:3)
```

Fancy plot via `?ggfortify:::autoplot.survfit`
```{r}
sf_1431 |> 
  autoplot() + xlim(0, 120) + labs(title = 'Figure 14.3.1')
```

Fancy plot via `?survminer:::ggsurvplot`
```{r}
sf_1431 |> 
  ggsurvplot(title = 'Figure 14.3.1')
```


### Example 14.4.1

Page 764 (10th ed), Page 659 (11th ed) 
```{r}
survdiff(OS ~ TUMOR, data = d1431)
```


### Example 14.5.1

Page 769 (10th ed), Page 663 (11th ed)  
```{r}
head(EXA_C14_S05_01)
head(d1451 <- within(EXA_C14_S05_01, expr = {
  time = .difftime(time, units = 'weeks')
  PFS = Surv(time, status)
  drug = relevel(structure(drug, levels = c('Opiate', 'Other'), class = 'factor'), ref = 'Other')
}))
```


```{r}
summary(model1_1451 <- coxph(PFS ~ drug + age, data = d1451))
```
'Opiate' has higher hazard compared to Drug='Other' (hazard ratio (HR) = 9.407, p < .001)
With 'proportional hazard' assumption, we don't need to discuss Opiate vs. Other at any specific time point

```{r}
confint(model1_1451)
```
'age' is not significant (p = .772), so it should be removed from the model

```{r}
summary(model2_1451 <- coxph(PFS ~ drug, data = d1451))
confint(model2_1451)
```
'Opiate' has higher hazard compared to Drug='Other' (hazard ratio (HR) = 9.923, p < .001)

plot
```{r}
sf_1451 = survfit(PFS ~ drug, data = d1451)
sf_1451 |> plot(lty = 2:3, xlab = 'weeks', ylab = 'PFS', main = 'Figure 14.5.1')
legend(x = 35, y = .9, legend = c('Other', 'Opiate'), lty = 2:3)
```

fancy plot
```{r}
sf_1451 |> autoplot() + labs(title = 'Figure 14.5.1')
```

another fancy plot
```{r}
sf_1451 |> ggsurvplot(title = 'Figure 14.5.1')
```


