---
title: "Wayne W. Daniel\\'s Biostatistics"
author: Tingting Zhan
format: 
  html:
    page-layout: full
    html-math-method: katex
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: visual
knitr:
  opts_chunk: 
    collapse: true
    comment: "#>" 
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette of package **`DanielBiostatistics10th`** ([CRAN](https://cran.r-project.org/package=DanielBiostatistics10th), [Github](https://github.com/tingtingzhan/DanielBiostatistics10th)) documents the examples of

-   ***Biostatistics: A Foundation for Analysis in the Health Sciences***, [10th Edition](https://dev.store.wiley.com/en-us/Biostatistics%3A+A+Foundation+for+Analysis+in+the+Health+Sciences%2C+10th+Edition-p-9781119625506)

by Wayne W. Daniel and Chad L. Cross.

## Note to Students

Examples in this vignette require that the `search` path has

```{r}
library(DanielBiostatistics10th)
library(flextable)
library(scales)
library(survival)
library(ggplot2)
library(ggfortify)
```

```{r}
#| echo: false
library(knitr) # for tables in this vignette
#options(mc.cores = 1L) # for CRAN submission
```

These are functions defined in package **`DanielBiostatistics10th`**, for each chapter of the book.

### Chapter 2

```{r}
#| echo: false
#| results: asis
c(
  '`?viewFreq`', 'View frequencies, cumulative frequencies and reversed cumulative frequencies of an `ordered` `factor`', 
  '`?viewStats`', 'View statistics (e.g., mean, standard deviations, etc.) of `numeric` observations'
) |>
  matrix(nrow = 2L, dimnames = list(c('Function', 'Description'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable() 
```

### Chapter 3

```{r}
#| echo: false
#| results: asis
c(
  '`?binTab`', '$2\\times 2$ Test-&-Disease or Risk-&-Disease table'
) |>
  matrix(nrow = 2L, dimnames = list(c('Function', 'Description'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable() 
```

### Chapter 5 to 7

```{r}
#| echo: false
#| results: asis
c(
  '`?z_test`', 'One- or two-sample $z$-test using aggregated statistics',
  '`?t_test`', 'One- or two-sample $t$-test using aggregated statistics',
  '`?prop_CLT`', 'One- or two-sample $z$-test on proportion(s) using aggregated statistics, based on Central Limit Theorem',
  '`?var_test`', 'One-sample $\\chi^2$-test or two-sample $F$-test on variances using aggregated statistics',
  '`?power_z`', 'Power curve of one-sample $z$-test.'
) |>
  matrix(nrow = 2L, dimnames = list(c('Function', 'Description'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable() 
```

### Chapter 12

```{r}
#| echo: false
#| results: asis
c(
  '`?viewOE`', 'View **o**bserved and **e**xpected frequencies of Pearson\'s $\\chi^2$-test'
) |>
  matrix(nrow = 2L, dimnames = list(c('Function', 'Description'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable() 
```

## Terms and Abbreviations

```{r}
#| echo: false
#| results: asis
c(
  '', 'Forward [pipe operator](https://search.r-project.org/R/refmans/base/html/pipeOp.html) introduced in `R` 4.1.0',
  '`$`', '[Extract](https://search.r-project.org/R/refmans/base/html/Extract.html) parts of an object',
  '[`abline`](https://search.r-project.org/R/refmans/graphics/html/abline.html)', 'Add straight line (by intercept and slope) to a plot',
  '[`aov`](https://search.r-project.org/R/refmans/stats/html/aov.html), [`anova`](https://search.r-project.org/R/refmans/stats/html/anova.html)', 'Analysis-of-variance (ANOVA) model; ANOVA table',
  '[`boxplot`](https://search.r-project.org/R/refmans/graphics/html/boxplot.html)', 'Box plot',
  '`CRAN`, `R`', '[The Comprehensive R Archive Network](https://cran.r-project.org)',
  '[`chisq.test`](https://search.r-project.org/R/refmans/stats/html/chisq.test.html)', 'Pearson\'s $\\chi^2$-test',
  '[`class`](https://search.r-project.org/R/refmans/base/html/class.html)', 'Object class',
  '[`confint`](https://search.r-project.org/R/refmans/stats/html/confint.html)', 'Confidence intervals for model parameters',
  '[`cut`](https://search.r-project.org/R/refmans/base/html/cut.html)', 'Convert numeric to factor',
  '[`data.frame`](https://search.r-project.org/R/refmans/base/html/data.frame.html)', 'Data frame',
  '[`dim`](https://search.r-project.org/R/refmans/base/html/dim.html)', 'Dimensions',
  '[`dbinom`, `pbinom`, `qbinom`](https://search.r-project.org/R/refmans/stats/html/Binomial.html)', 'Binomial density, probability, quantile',
  '[`dpois`, `ppois`, `qpois`](https://search.r-project.org/R/refmans/stats/html/Poisson.html)', 'Poisson density, probability, quantile',
  '[`dnorm`, `pnorm`, `qnorm`](https://search.r-project.org/R/refmans/stats/html/Normal.html)', 'Normal density, probability, quantile',
  '[`exp`](https://search.r-project.org/R/refmans/base/html/Log.html)', 'Exponential',
  '[`factor`](https://search.r-project.org/R/refmans/base/html/factor.html)', 'Factor',
  '[`fisher.test`](https://search.r-project.org/R/refmans/stats/html/fisher.test.html)', 'Fisher\'s exact test', 
  '[`flextable`](https://search.r-project.org/CRAN/refmans/flextable/html/flextable.html)', 'Flexible tables',
  '[`glm`](https://search.r-project.org/R/refmans/stats/html/glm.html)', 'Generalized linear model',
  '[`head`, `tail`](https://search.r-project.org/R/refmans/utils/html/head.html)', 'First or last parts of an object',
  '[`kurtosis`](https://search.r-project.org/CRAN/refmans/e1071/html/kurtosis.html)', 'Kurtosis',
  '[`L`](https://adv-r.hadley.nz/vectors-chap.html#atomic-vectors)', '[Integer](https://search.r-project.org/R/refmans/base/html/integer.html) notation',
  '[`label_percent`](https://search.r-project.org/CRAN/refmans/scales/html/label_percent.html)', 'Label for percentages',
  '[`lm`](https://search.r-project.org/R/refmans/stats/html/lm.html)', 'Linear model',
  '[`log`](https://search.r-project.org/R/refmans/base/html/Log.html)', 'Logarithm',
  '[`matrix`](https://search.r-project.org/R/refmans/base/html/matrix.html)', 'Matrices',
  '[`mean`](https://search.r-project.org/R/refmans/base/html/mean.html)', 'Arithmetic mean',
  '[`median`](https://search.r-project.org/R/refmans/stats/html/median.html)', 'Median',
  '[`Mode`](https://search.r-project.org/CRAN/refmans/pracma/html/mode.html)', 'Mode',
  '[`names`](https://search.r-project.org/R/refmans/base/html/names.html)', 'Names of an object',
  '[`nrow`](https://search.r-project.org/R/refmans/base/html/nrow.html)', 'Number of rows',
  '[`ordered`](https://search.r-project.org/R/refmans/base/html/factor.html)', 'Ordered factor',
  '[`plot`](https://search.r-project.org/R/refmans/base/html/plot.html)', 'Plot',
  '[`prop.test`](https://search.r-project.org/R/refmans/stats/html/prop.test.html)', 'Pearson\'s $\\chi^2$-test on proportions',
  '[`quantile`](https://search.r-project.org/R/refmans/stats/html/quantile.html)', 'Quantile',
  '[`rep`](https://search.r-project.org/R/refmans/base/html/rep.html)', 'Replicate elements',
  '[`sample`, `sample.int`](https://search.r-project.org/R/refmans/base/html/sample.html)', 'Random samples', 
  '[`sd`](https://search.r-project.org/R/refmans/stats/html/sd.html)', 'Standard deviation',
  '[`search`](https://search.r-project.org/R/refmans/base/html/search.html)', 'Search path',
  '[`seed`, `set.seed`](https://search.r-project.org/R/refmans/base/html/Random.html)', 'Random number generation seed',
  '[`seq`](https://search.r-project.org/R/refmans/base/html/seq.html)', 'Sequence generation',
  '[`skewness`](https://search.r-project.org/CRAN/refmans/e1071/html/skewness.html)', 'Skewness',
  '[`sort`](https://search.r-project.org/R/refmans/base/html/sort.html)', 'Sorting vectors',
  '[`summary`](https://search.r-project.org/R/refmans/base/html/summary.html)', 'Summaries of an `R` object',
  '[`t.test`](https://search.r-project.org/R/refmans/stats/html/t.test.html)', 'Student\'s $t$-test',
  '[`table`](https://search.r-project.org/R/refmans/base/html/table.html)', 'Cross tabulation',
  '[`uniroot`](https://search.r-project.org/R/refmans/stats/html/uniroot.html)', 'One dimensional root-finding',
  '[`var`](https://search.r-project.org/R/refmans/stats/html/cor.html)', 'Variance',
  '[`var.test`](https://search.r-project.org/R/refmans/stats/html/var.test.html)', '$F$-test to compare variances',
  '[`with`, `within`](https://search.r-project.org/R/refmans/base/html/with.html)', 'Evaluate an expression in a data environment', 
  '[`xtabs`](https://search.r-project.org/R/refmans/stats/html/xtabs.html)', 'Cross tabulation'
) |>
  matrix(nrow = 2L, dimnames = list(c('Term / Abbreviation', 'Description'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable() 
```

# Chapter 1: *Introduction to Biostatistics*

## Section 1.4 *Sampling and Statistical Inference*

### Example 1.4.1

Page 8 (10th ed), Page 7 (11th ed).

```{r}
EXA_C01_S04_01 |> class() 
```

```{r}
EXA_C01_S04_01 |> head(n = 3L)
```

```{r}
set.seed(12); EXA_C01_S04_01[sample.int(n = nrow(EXA_C01_S04_01), size = 10L), ] |> 
  sort_by(y = ~ SUBJ)
```

### Example 1.4.2

Page 11 (10th ed), Page 10 (11th ed).

```{r}
EXA_C01_S04_01[seq(from = 4L, to = 166L, by = 18L), ]
```

# Chapter 2: *Descriptive Statistics*

## Section 2.2 *The Ordered Array*

### Example 2.2.1

Page 20 (10th ed), Page 19 (11th ed)

```{r}
EXA_C01_S04_01$AGE |> class()
```

```{r}
EXA_C01_S04_01$AGE |> head()
```

```{r}
EXA_C01_S04_01$AGE |> sort() |> head()
```

## Section 2.3 *Grouped data: the Frequency Distribution*

### Example 2.3.1

Page 23 (10th ed); Page 22 (11th ed).

```{r}
EXA_C01_S04_01$AGE |>
  cut(breaks = seq(from = 30, to = 90, by = 10), right = FALSE, ordered_result = TRUE) |>
  viewFreq()
```

### Exercise 2.3.1

Page 30

```{r}
EXR_C02_S03_01$pindex |> viewStats()
```

```{r}
EXR_C02_S03_01$pindex |>
  cut(breaks = seq(0, 3, by = .5), right = FALSE, ordered_result = TRUE) |>
  viewFreq()
```

### Exercise 2.3.2

Page 31

```{r}
EXR_C02_S03_02$sizes |>
  cut(breaks = seq(0, 30, by = 5), right = FALSE, ordered_result = TRUE) |>
  viewFreq()
```

### Exercise 2.3.3

Page 32

```{r}
EXR_C02_S03_03$bmi |> viewStats()
```

```{r}
EXR_C02_S03_03$bmi |>
  cut(breaks = seq(20, 55, by = 5), right = FALSE, ordered_result = TRUE) |>
  viewFreq()
```

### Exercise 2.3.11

Page 36

```{r}
EXR_C02_S03_11$SRratio |> viewStats()
```

```{r}
EXR_C02_S03_11$SRratio |>
  cut(breaks = seq(0, 1.2, by = .1), right = TRUE, ordered_result = TRUE) |>
  viewFreq()
```

```{r}
EXR_C02_S03_11$SRratio |>
  cut(breaks = seq(0, 1.2, by = .1), right = FALSE, ordered_result = TRUE) |>
  viewFreq()
```

## Section 2.4 *Descriptive Statistics: Measures of Central Tendency*

### Example 2.4.1-2.4.6

Page 38-42 (10th ed); Page 30-34 (11th ed).

```{r}
EXA_C01_S04_01$AGE |> viewStats()
```

## Section 2.5 *Descriptive Statistics: Measures of Dispersion*

### Example 2.5.1-2.5.3

Page 44-46 (10th ed); Page 35-41 (11th ed).

See previous section.

### Example 2.5.4

Page 49 (10th ed), Page 41 (11th ed)

Data not available. Omitted.

### Example 2.5.5

Page 50 (10th ed)

```{r}
#| fig-width: 4
EXA_C02_S05_05$GRF |> boxplot(main = 'EXA_C02_S05_05$GRF')
```

```{r}
EXA_C02_S05_05$GRF |> viewStats()
```

## Large Data Set

Page 62,

```{r}
LDS_C02_NCBIRTH800 |> head(n = 3L)
```

```{r}
NCBIRTH = within(LDS_C02_NCBIRTH800, expr = {
  sex = structure(sex, levels = c('male', 'female'), class = 'factor')
  marital = structure(marital, levels = c('married', 'not married'), class = 'factor')
  smoke = structure(smoke + 1L, levels = c('not smoked', 'smoked'), class = 'factor')
  premie = structure(premie + 1L, levels = c('not premature', 'premature'), class = 'factor')
  low = structure(low + 1L, levels = c('not low', 'low'), class = 'factor')
  racemom = structure(racemom + 1L, levels = c('Other_non_white', 'White', 'Black', 'Indian', 'Chinese', 'Japanese', 'Hawaiian', 'Filipino', 'Other_Asian_PacificIs'), class = 'factor')
})
NCBIRTH |> head(n = 3L)
```

The benefit of using `factor` instead of `integer`.

```{r}
LDS_C02_NCBIRTH800$sex |> table()
NCBIRTH$sex |> table()
```

```{r}
NCBIRTH$mage |> viewStats()
```

```{r}
NCBIRTH$weeks |> viewStats()
```

```{r}
NCBIRTH$gained |> viewStats()
```

```{r}
NCBIRTH$tounces |> viewStats()
```

```{r}
NCBIRTH$tgrams |> viewStats()
```

```{r}
range(NCBIRTH$tounces / NCBIRTH$tgrams)
NCBIRTH |> 
  with(expr = tounces / tgrams) |> 
  range()
```

```{r}
NCBIRTH$mage |> boxplot(main = 'mage')
```

```{r}
NCBIRTH$weeks |> boxplot(main = 'weeks')
```

```{r}
NCBIRTH$gained |> boxplot(main = 'gained')
```

```{r}
NCBIRTH$tounces |> boxplot(main = 'tounces')
```

```{r}
boxplot(tounces ~ smoke, data = NCBIRTH)
```

```{r}
boxplot(mage ~ marital, data = NCBIRTH)
```

# Chapter 3: *Some Basic Probability Concepts*

## Section 3.4 *Calculating the Probability of an Event*

### Example 3.4.1 - 3.4.9

Page 69-75 (10th ed), Page 61-67 (11th ed)

```{r}
c(28L, 19L, 41L, 53L, 35L, 38L, 44L, 60L) |>
  matrix(ncol = 2L, dimnames = list(
    FamilyHx = c('none', 'Bipolar', 'Unipolar', 'UniBipolar'), 
    Onset = c('Early', 'Late')
  )) |> 
  as.table() |>
  as_flextable()
```

## Section 3.5 *Bayes' Theorem, Screening Tests, Sensitivity, Specificity, and Predictive Value Positive and Negative*

### Example 3.5.1

Page 81 (10th ed), Page 72 (11th ed)

```{r}
d351 = c(495L, 14L, 5L, 436L) |>
  matrix(nrow = 2L, dimnames = list(
    Alzheimer = c('No', 'Yes'), 
    Test = c('Negative', 'Positive')
  )) |>
  binTab()
```

```{r}
d351 |> as_flextable()
```

```{r}
d351 |> summary(prevalence = .113)
```

### Exercise 3.5.1

Page 83 (10th ed)

```{r}
e351 = c(1380L - 21L, 775L - 744L, 21L, 744L) |> 
  matrix(ncol = 2L) |>
  binTab(nm = c('Disease', 'Symptom'))
e351 |> as_flextable()
e351 |> summary(prevalence = c(.0001, .001, .01, .1))
```

### Exercise 3.5.2

Page 83 (10th ed)

```{r}
e352 = c(18L, 5L, 10L, 38L) |>
  matrix(ncol = 2L) |>
  binTab(nm = c('Confirmed Tear', 'Absent Bow-Tie'))
e352 |> as_flextable()
e351 |> summary()
```

### Exercise 3.5.3

Page 83 (10th ed)

```{r}
ppv(prevalence = 1/32000, sensitivity = .927, specificity = .997)
```

```{r}
npv(prevalence = 1/32000, sensitivity = .927, specificity = .997)
```

## Review

### Review 3

Page 85 (10th ed)

```{r}
c(319L, 130L, 88L, 41L, 738L, 329L, 402L, 95L) |>
  matrix(nrow = 4L, dimnames = list(
    MaritalStatus = c('Currently Married', 'Divorced', 'Widowed', 'Unmarried'),
    Ethnicity = c('Hispanic', 'Non-Hispanic'))) |>
  as.table() |>
  as_flextable()
```

### Review 5

Page 87 (10th ed)

```{r}
c(9L, 20L, 5L, 4L, 4L, 2L, 17L, 3L, 7L, 13L, 6L, 15L, 8L, 4L, 9L) |>
  matrix(nrow = 5L, dimnames = list(
    Personality = c('Obsessoid', 'Asthenic / Low Self-Confident', 'Asthenic / High Self-Confident', 'Nervous / Tense', 'Undeterminable'), 
    Group = c('ATPD', 'PS', 'BSAD'))) |>
  as.table() |> 
  as_flextable()
```

### Review 9

Page 88 (10th ed)

```{r}
r39 = c(115849L, 112L, 2244L, 2991L) |>
  matrix(nrow = 2L, dimnames = list(
    'CancerInRegistry' = c('No', 'Yes'),
    'CancerReported' = c('No', 'Yes'))) |>
  binTab()
r39 |> as_flextable()
```

```{r}
r39 |> summary()
```

## Large Data Set

Page 90

```{r}
xtabs(~ premie + smoke, data = NCBIRTH) |> as_flextable()
```

```{r}
xtabs(~ low + marital, data = NCBIRTH) |> as_flextable()
```

# Chapter 4: *Probability Distributions*

## Section 4.2 *Probability Distributions of Discrete Variables*

### Example 4.2.1-4.2.7

Page 93-97 (10th ed), Page 81-85 (11th ed)

```{r}
(d421 = rep(1:8, times = c(62L, 47L, 39L, 39L, 58L, 37L, 4L, 11L)))
```

Table 4.2.1, 4.2.2, Table 4.2.3

```{r}
d421 |>
  ordered() |>
  viewFreq()
```

### Example 4.2.8

Page 98 (10th ed), Page 85 (11th ed).

```{r}
d421 |> viewStats()
```

## Section 4.3 *The Binomial Distribution*

### Example 4.3.1

Page 99 (10th ed).

`d` for `density`; calculates $\text{Pr}(X = x)$.

```{r}
dbinom(x = 3L, size = 5L, prob = .858) |> 
  label_percent(accuracy = .1)()
```

### Example 4.3.1

Page 87 (11th ed).

```{r}
dbinom(x = 3L, size = 5L, prob = .899) |> 
  label_percent(accuracy = .1)()
```

### Example 4.3.2

Page 103 (10th ed), Page 90 (11th ed).

```{r}
dbinom(x = 4L, size = 10L, prob = .14) |> 
  label_percent(accuracy = .1)()
```

### Example 4.3.3

Page 103 (10th ed), Page 91 (11th ed).

`p` for `probability`; calculates

-   $\text{Pr}(X \leq x)$ if `lower.tail = TRUE` (default)
-   $\text{Pr}(X > x)$ if `lower.tail = FALSE`

```{r}
pbinom(q = 5L, size = 25L, prob = .1, lower.tail = TRUE) |>
  label_percent(accuracy = .1)()
pbinom(q = 5L, size = 25L, prob = .1, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
```

### Example 4.3.4

Page 105 (10th ed), Page 92 (11th ed)

#### (a)

```{r}
dbinom(x = 7L, size = 12L, prob = .55) |> 
  label_percent(accuracy = .1)()
```

#### (b)

```{r}
pbinom(q = 5L, size = 12L, prob = .55) |> 
  label_percent(accuracy = .1)()
```

#### (c)

```{r}
pbinom(q = 7L, size = 12L, prob = .55, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
```

## Section 4.4 *The Poission Distribution*

### Example 4.4.1

Page 110 (10th ed), Page 97 (11th ed)

```{r}
dpois(x = 3L, lambda = 12) |> 
  label_percent(accuracy = .1)()
```

### Example 4.4.2

Page 110 (10th ed), Page 98 (11th ed)

```{r}
ppois(q = 2L, lambda = 12, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
```

### Example 4.4.3

Page 110 (10th ed), Page 98 (11th ed)

```{r}
ppois(q = 1L, lambda = 2) |> 
  label_percent(accuracy = .1)()
```

### Example 4.4.4

Page 111 (10th ed), Page 98 (11th ed)

```{r}
dpois(x = 3L, lambda = 2) |> 
  label_percent(accuracy = .1)()
```

### Example 4.4.5

Page 112 (10th ed), Page 98 (11th ed)

```{r}
ppois(q = 5L, lambda = 2, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
```

## Section 4.6 *The Normal Distribution*

### Example 4.6.1

Page 119 (10th ed), Page 106 (11th ed)

```{r}
pnorm(q = 2) |> 
  label_percent(accuracy = .1)()
```

### Example 4.6.2

Page 120 (10th ed), Page 106 (11th ed)

```{r}
(pnorm(2.55) - pnorm(-2.55)) |> 
  label_percent(accuracy = .1)()
```

```{r}
(1 - 2 * pnorm(-2.55)) |> 
  label_percent(accuracy = .1)() # alternative solution
```

### Example 4.6.3

Page 121 (10th ed), Page 107 (11th ed)

```{r}
(pnorm(1.53) - pnorm(-2.74)) |> 
  label_percent(accuracy = .1)()
```

### Example 4.6.4

Page 121 (10th ed), Page 107 (11th ed)

```{r}
pnorm(2.71, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
```

### Example 4.6.5

Page 122 (10th ed), Page 107 (11th ed)

```{r}
(pnorm(2.45) - pnorm(.84)) |> 
  label_percent(accuracy = .1)()
```

## Section 4.7 *Normal Distribution Applications*

### Example 4.7.1

Page 122 (10th ed), Page 109 (11th ed)

```{r}
pnorm(q = 3, mean = 5.4, sd = 1.3) |> 
  label_percent(accuracy = .1)()
```

### Example 4.7.2

Page 125 (10th ed), Page 111 (11th ed)

```{r}
(pnorm(q = 649, mean = 491, sd = 119) - pnorm(q = 292, mean = 491, sd = 119)) |> 
  label_percent(accuracy = .1)()
```

### Example 4.7.3

Page 122 (10th ed), Page 111 (11th ed)

```{r}
pnorm(q = 8.5, mean = 5.4, sd = 1.3, lower.tail = FALSE) |> 
  label_percent(accuracy = .01)()
```

## Exercise

### Exercise 4.4.4

Page 113,

```{r}
dpois(1L, lambda = .5) |> 
  label_percent(accuracy = .01)()
```

```{r}
dpois(0L, lambda = .5) |> 
  label_percent(accuracy = .01)()
```

```{r}
dpois(4L, lambda = .5) |> 
  label_percent(accuracy = .01)()
```

```{r}
ppois(0L, lambda = .5, lower.tail = FALSE) |> 
  label_percent(accuracy = .01)()
```

### Exercise 4.6.11

Page 122

```{r}
qnorm(.0055)
```

### Exercise 4.6.12

Page 122.

P(-2.67\<=Z\<=z) = .9718

```{r}
# pnorm(z) - pnorm(-2.67) = .9718
qnorm(.9718 + pnorm(-2.67))
```

## Review

### Review 15

Page 130

```{r}
dbinom(x = 7L, size = 10L, prob = .35) |> 
  label_percent(accuracy = .1)()
pbinom(q = 5L, size = 10L, prob = .35, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
dbinom(x = 0L, size = 10L, prob = .35) |> 
  label_percent(accuracy = .1)()
(pbinom(q = 6L, size = 10L, prob = .35) - pbinom(q = 2L, size = 10L, prob = .35)) |> 
  label_percent(accuracy = .1)()
```

### Review 17

Page 130,

```{r}
dpois(10L, lambda = 15.8) |> 
  label_percent(accuracy = .1)()
(ppois(15L, lambda = 15.8) - ppois(4L, lambda = 15.8)) |> 
  label_percent(accuracy = .1)()
ppois(10L, lambda = 15.8, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
ppois(7L, lambda = 15.8) |> 
  label_percent(accuracy = .1)()
(ppois(11L, lambda = 15.8) - ppois(8L, lambda = 15.8)) |> 
  label_percent(accuracy = .1)()
```

### Review 19

Page 131,

```{r}
pbinom(6L, size = 8L, prob = .8) |> 
  label_percent(accuracy = .1)()
pbinom(6L, size = 8L, prob = .8, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
dbinom(8L, size = 8L, prob = .8) |> 
  label_percent(accuracy = .1)()
pbinom(3L, size = 8L, prob = .8) |> 
  label_percent(accuracy = .1)()
(pbinom(7L, size = 8L, prob = .8) - pbinom(3L, size = 8L, prob = .8)) |> 
  label_percent(accuracy = .1)()
```

### Review 20

Page 131 (10th ed)

```{r}
dpois(0L, lambda = 1.3) |> 
  label_percent(accuracy = .1)()
ppois(0L, lambda = 1.3, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
ppois(4L, lambda = 1.3) |> 
  label_percent(accuracy = .1)()
```

### Review 23

Page 131 (10th ed),

```{r}
pnorm(200, mean = 500, sd = sqrt(1e4)) |> 
  label_percent(accuracy = .1)()
pnorm(650, mean = 500, sd = sqrt(1e4), lower.tail = FALSE) |> 
  label_percent(accuracy = .1)()
(pnorm(675, mean = 500, sd = sqrt(1e4)) - pnorm(350, mean = 500, sd = sqrt(1e4))) |> 
  label_percent(accuracy = .1)()
```

### Review 26

Page 132 (10th ed),

```{r}
(rev426 = qnorm((1-.754)/2, lower.tail = FALSE))
(pnorm(rev426) - pnorm(-rev426)) |> 
  label_percent(accuracy = .1)()
```

### Review 27

Page 132 (10th ed)

```{r}
qnorm(.0094, mean = 100, sd = 15, lower.tail = TRUE)
qnorm(.1093, mean = 100, sd = 15, lower.tail = FALSE)
qnorm(.5+.4778, mean = 100, sd = 15, lower.tail = TRUE) 
qnorm(.9660+((1-.9660)/2), mean = 100, sd = 15, lower.tail = TRUE) 
```

### Review 28

Page 132 (10th ed)

```{r}
# (40-mu)/sigma = qnorm(.0080)
(mu428 = 40 - qnorm(.0080) * 10)
qnorm(.0080, mean = mu428, sd = 10, lower.tail = TRUE)
```

# Chapter 5, *Some Important Sampling Distributions*

## Section 5.3

### Example 5.3.2

Page 142 (10th ed)

```{r}
z_test(xbar = 190, sd = 12.7, n = 10L, null.value = 185.6, alternative = 'greater')
```

### Example 5.3.3

Page 143 (10th ed)

```{r}
(pnorm(125, mean = 120, sd = 15/sqrt(50)) - pnorm(115, mean = 120, sd = 15/sqrt(50))) |> 
  label_percent(accuracy = .1)()
(z_test(xbar = 125, sd = 15, n = 50L, null.value = 120, alternative = 'less')$p.value - 
  z_test(xbar = 115, sd = 15, n = 50L, null.value = 120, alternative = 'less')$p.value) |> 
  label_percent(accuracy = .1)()
```

Page 127 (11th ed)

```{r}
(z_test(xbar = 110, sd = 20, n = 50L, null.value = 100, alternative = 'less')$p.value - 
  z_test(xbar = 95, sd = 20, n = 50L, null.value = 100, alternative = 'less')$p.value) |> 
  label_percent(accuracy = .1)()
```

### Exercise 5.3.7

Page 145,

```{r}
(pnorm(q = 55, mean = 50, sd = 16/sqrt(64L)) - pnorm(q = 45, mean = 50, sd = 16/sqrt(64L))) |> 
  label_percent(accuracy = .1)()
```

```{r}
pnorm(q = 53, mean = 50, sd = 16/sqrt(64L), lower.tail = FALSE)  |> 
  label_percent(accuracy = .1)()
```

## Section 5.4

### Example 5.4.1

Page 145 (10th ed), Page 129 (11th ed)

```{r}
z_test(xbar = c(92, 105), sd = 20, n = 15L, null.value = 0, alternative = 'less') 
```

### Example 5.4.2

Page 148 (10th ed),

```{r}
z_test(xbar = 20, sd = c(15, 20), n = c(35L, 40L), null.value = c(45, 30), alternative = 'greater')
```

## Section 5.5

### Example 5.5.1

Page 150 (10th ed),

```{r}
prop_CLT(phat = .4, n = 150L, null.value = .357, alternative = 'greater')
```

### Example 5.5.2

Page 152 (10th ed),

```{r}
prop_CLT(phat = .45, n = 200L, null.value = .51, alternative = 'less')
```

## Section 5.6

### Example 5.6.1

Page 155 (10th ed),

```{r}
prop_CLT(phat = .1, null.value = c(.28, .21), n = c(100L, 100L), alternative = 'greater')
```

### Example 5.6.2

Page 155 (10th ed),

```{r}
prop_CLT(phat = .05, null.value = c(.34, .26), n = c(250L, 200L), alternative = 'less')
```

## Review

### Review 11

Page 158 (10th ed).

```{r}
pnorm(25, mean = 23.1, sd = 3.7/sqrt(45L), lower.tail = FALSE)
```

```{r}
z_test(xbar = 25, n = 45L, sd = 3.7, null.value = 23.1, alternative = 'greater')
```

### Review 12

Page 158 (10th ed).

```{r}
pnorm(24, mean = 24.7, sd = 3.3/sqrt(50L), lower.tail = TRUE) |> 
  label_percent(accuracy = .1)()
```

```{r}
z_test(xbar = 24, n = 50L, sd = 3.3, null.value = 24.7, alternative = 'less')
```

### Review 13

Page 158 (10th ed).

```{r}
z_test(xbar = 3, n = c(45L, 50L), sd = c(3.7, 3.3), null.value = c(23.1, 24.7), alternative = 'greater')
```

### Review 14

Page 158 (10th ed)

### Review 17

Page 159,

```{r}
prop_CLT(phat = .25, n = 65L, null.value = .19, alternative = 'greater')
```

### Review 18

Page 159,

```{r}
prop_CLT(phat = .2, n = 220L, null.value = .17, alternative = 'greater')
```

### Review 19

Page 159,

```{r}
prop_CLT(phat = .2, n = 250L, null.value = .23, alternative = 'less')
```

### Review 20

Page 159

```{r}
prop_CLT(phat = .05, n = c(250L, 220L), null.value = c(.23, .17), alternative = 'less')
```

### Review 21

(not done)

# Chapter 6, *Estimation*

## Section 6.2

### Example 6.2.1

Page 166 (10th ed), Page 147 (11th ed)

```{r}
z_test(xbar = 22, n = 10L, sd = sqrt(45))
```

### Example 6.2.2

Page 168 (10th ed), Page 149 (11th ed)

```{r}
z_test(xbar = 84.3, n = 15L, sd = sqrt(144), conf.level = .99)
```

### Example 6.2.3

Page 168 (10th ed), Page 150 (11th ed)

```{r}
z_test(xbar = 17.2, n = 35L, sd = 8, conf.level = .9)
```

### Example 6.2.4

Page 169 (10th ed), Page 150 (11th ed)

```{r}
z_test(obs = EXA_C06_S02_04$ACTIVITY, sd = sqrt(.36))
```

### Exercise 6.2.1

Page 171,

```{r}
z_test(xbar = 90, sd = 10, n = 49L, conf.level = .9)
```

```{r}
z_test(xbar = 90, sd = 10, n = 49L, conf.level = .95)
```

```{r}
z_test(xbar = 90, sd = 10, n = 49L, conf.level = .99)
```

## Section 6.3 The $t$-Distribution

### Example 6.3.1

Page 173 (10th ed),

```{r}
t_test(xbar = 250.8, xsd = 130.9, n = 19L)
```

## Section 6.4

### Example 6.4.1

Page 177 (10th ed),

```{r}
z_test(xbar = c(4.5, 3.4), sd = sqrt(c(1, 1.5)), n = c(12L, 15L))
```

### Example 6.4.2

Page 178 (10th ed),

```{r}
z_test(xbar = c(4.3, 13), sd = c(5.22, 8.97), n = c(328L, 64L), conf.level = .99)
```

### Example 6.4.3

Page 180 (10th ed),

```{r}
t_test(xbar = c(4.7, 8.8), xsd = c(9.3, 11.5), n = c(18L, 10L), var.equal = TRUE)
```

### Example 6.4.4

Page 181 (10th ed),

```{r}
t_test(xbar = c(4.7, 8.8), xsd = c(9.3, 11.5), n = c(18L, 10L))
```

Welch slightly different from Cochran; textbook explained on Page 182

## Section 6.5

### Example 6.5.1

Page 185 (10th ed),

```{r}
prop_CLT(phat = .18, n = 1220L)
```

## Section 6.6

### Example 6.6.1

Page 187 (10th ed),

```{r}
prop_CLT(x = c(31L, 53L), n = c(68L, 255L), conf.level = .99)
```

```{r}
prop.test(x = c(31L, 53L), n = c(68L, 255L), conf.level = .99)
```

## Section 6.7

### Example 6.7.1

Page 190 (10th ed),

```{r}
uniroot(f = function(n, sd, level = .95) {
  qnorm(1-(1-level)/2) * sd/sqrt(n) - 5 # half-width of CI <= 5 grams
}, interval = c(0, 2e2), sd = 20)$root |>
  ceiling()
```

## Section 6.8

### Example 6.8.1

Page 192 (10th ed),

```{r}
uniroot(f = function(n, p, level = .95) {
  qnorm(1-(1-level)/2) * sqrt(p*(1-p)/n) - .05
}, interval = c(0, 1e3), p = .35)$root |>
  ceiling()
```

## Section 6.9

### Example 6.9.1

Page 196 (10th ed),

```{r}
d691 = c(9.7, 12.3, 11.2, 5.1, 24.8, 14.8, 17.7)
var_test(obs = d691) |> sqrt()
```

## Section 6.10

### Example 6.10.1

Page 200 (10th ed),

```{r}
var_test(xsd = c(8.1, 5.9), n = c(16L, 4L))
```

## Review

### Review 18

Page 206,

```{r}
t_test(xbar = c(3.85, 2.8), n = c(21L, 16L), xsd = c(1.95, 1.7), conf.level = .95)
```

```{r}
t_test(xbar = c(3.85, 2.8), n = c(21L, 16L), xsd = c(1.95, 1.7), conf.level = .95, var.equal = TRUE)
```

### Review 20

Page 206,

```{r}
t_test(xbar = c(95, 125), n = c(35L, 35L), xsd = c(25, 30), conf.level = .95)
```

```{r}
t_test(xbar = c(95, 125), n = c(35L, 35L), xsd = c(25, 30), conf.level = .95, var.equal = TRUE)
```

### Review 23

Page 207,

```{r}
REV_C06_23 |> head(n = 3L)
```

```{r}
t.test(Hours ~ Group, data = REV_C06_23, var.equal = TRUE) 
```

### Review 24

Page 207,

```{r}
z_test(xbar = c(358, 483), sd = c(308, 144), n = c(52L, 53L), conf.level = .99)
```

### Review 25

Page 207,

```{r}
t_test(xbar = 435, n = 34L, xsd = 215, conf.level = .95)
```

### Review 26

Page 207,

```{r}
prop_CLT(phat = c(.293, .173), n = c(220L, 106L), conf.level = .95)
```

### Review 28

Page 208,

```{r}
r628 = REV_C06_28 |>
  within(expr = {
    Group = structure(Group, levels = c('MyocardialInfarction', 'Healthy'), class = 'factor')
  })
```

```{r}
t.test(LA ~ Group, data = r628) 
```

## Large Data Sets

Page 210,

### Question 1

#### 1(a)

```{r}
NCBIRTH |> with(expr = prop_CLT(obs = (sex == 'male')))
```

#### 1(b)

```{r}
NCBIRTH$mage |> t.test()
```

#### 1(c)

```{r}
NCBIRTH$gained |> t.test()
```

#### 1(d)

```{r}
NCBIRTH |> with(expr = prop_CLT(obs = (smoke == 'smoked')))
```

#### 1(e)

```{r}
t.test(gained ~ smoke, data = NCBIRTH)
```

#### 1(f)

```{r}
t.test(tounces ~ marital, data = NCBIRTH)
```

#### 1(g)

```{r}
prop_CLT(x = 35L, n = c(263L, 537L))
```

```{r}
prop.test(x = rep(35L, times = 2L), n = c(263L, 537L))
```

### Question 2

```{r}
LDS_C06_CHOLEST$CHOLEST |> 
  sample(size = 15L) |> 
  t.test()
```

### Question 3

```{r}
prop_CLT(obs = (sample(LDS_C06_CHOLEST$CHOLEST, size = 50L) > 225))
```

### Question 4

```{r}
LDS_C06_BABYWGTS$WGT |> 
  sample(size = 20L) |>
  t.test()
```

### Question 5

```{r}
LDS_C06_BABYWGTS$WGT |>
  sample(size = 35L) |>
  t.test() 
```

### Question 6

```{r}
LDS_C06_BOYHGTS$HGT |>
  sample(size = 15L) |>
  t.test(conf.level = .99)
```

### Question 7

```{r}
LDS_C06_BOYHGTS$HGT |>
  sample(size = 35L) |>
  t.test(conf.level = .99)
```

# Chapter 7: *Hypothesis Testing*

## Section 7.2

### Example 7.2.1

Page 222 (10th ed); Page 201 (11th ed)

```{r}
z_test(xbar = 27, sd = sqrt(20), n = 10L, null.value = 30)
```

### Example 7.2.2

Page 226 (10th ed); Page 204 (11th ed)

```{r}
z_test(xbar = 27, sd = sqrt(20), n = 10L, null.value = 30, alternative = 'less')
```

### Example 7.2.3

Page 228 (10th ed); Page 206 (11th ed)

```{r}
EXA_C07_S02_03$DAYS |> t.test(mu = 15)
```

### Example 7.2.4

Page 231 (10th ed); Page 209 (11th ed)

```{r}
z_test(xbar = 146, sd = 27, n = 157L, null.value = 140, alternative = 'greater')
```

### Example 7.2.5

Page 232 (10th ed); Page 210 (11th ed)

```{r}
c(33.38, 32.15, 34.34, 33.95, 33.46, 34.13, 33.99, 34.10, 33.85, 34.23, 34.45, 34.19, 33.97, 32.73, 34.05) |> 
  t.test(mu = 34.5)
```

### Exercise 7.2.15

Page 235

```{r}
EXR_C07_S02_15$Age |> t.test(mu = 30, alternative = 'less')
```

## Section 7.3

### Example 7.3.1

Page 237 (10th ed), Page 213 (11th ed)

```{r}
z_test(xbar = c(4.5, 3.4), sd = sqrt(c(1, 1.5)), n = c(12L, 15L), null.value = 0)
```

### Example 7.3.2

Page 239 (10th ed), Page 215 (11th ed)

```{r}
EXA_C07_S03_02 |>
  with(expr = t.test(x = CONTROL, y = SCI, alternative = 'less', var.equal = TRUE))
```

### Example 7.3.3

Page 240 (10th ed), Page 217 (11th ed)

```{r}
t_test(xbar = c(19.16, 9.53), xsd = c(5.29, 2.69), n = c(15L, 30L), null.value = 0)
```

### Example 7.3.4

Page 242 (10th ed), Page 219 (11th ed)

```{r}
z_test(xbar = c(59.01, 46.61), sd = c(44.89, 34.85), n = c(53L, 54L), null.value = 0,
  alternative = 'greater')
```

### Exercise 7.3.4

Page 246

```{r}
e734 = EXR_C07_S03_04 |>
  within(expr = {
    Group <- structure(Group, levels = c('PH', 'NT'), class = 'factor')
  }) 
```

```{r}
t.test(Length ~ Group, data = e734, alternative = 'greater')
```

## Section 7.4

### Example 7.4.1

Page 251 (10th ed), Page 226 (11th ed)

```{r}
EXA_C07_S04_01 |>
  with(expr = t.test(x = POSTOP, y = PREOP, alternative = 'greater', paired = TRUE))
```

## Section 7.5

### Example 7.5.1

Page 258 (10th ed), Page 232 (11th ed)

```{r}
prop_CLT(x = 24L, n = 301L, null.value = .063, alternative = 'greater') 
```

```{r}
prop.test(x = 24L, n = 301L, p = .063, alternative = 'greater')
```

### Exercise 7.5.4

Page 260,

```{r}
prop_CLT(x = 175L, n = 250L, null.value = .6, alternative = 'greater', conf.level = .99)
```

```{r}
prop.test(x = 175L, n = 250L, p = .6, alternative = 'greater', conf.level = .99)
```

### Exercise 7.5.5

Page 260

```{r}
prop_CLT(x = 25L, n = 250L, null.value = .15, alternative = 'less')
```

```{r}
prop.test(x = 25L, n = 250L, p = .15, alternative = 'less')
```

### Exercise 7.5.6

Page 260

```{r}
prop_CLT(x = 238L, n = 250L, null.value = .9, alternative = 'greater')
```

```{r}
prop.test(x = 238L, n = 250L, p = .9, alternative = 'greater')
```

## Section 7.6

### Example 7.6.1

Page 261 (10th ed), Page 235 (11th ed)

```{r}
prop_CLT(x = c(24L, 11L), n = c(44L, 29L), null.value = 0, alternative = 'greater')
```

```{r}
prop.test(x = c(24L, 11L), n = c(44L, 29L), alternative = 'greater')
```

## Section 7.7

### Example 7.7.1

Page 264 (10th ed), Page 238 (11th ed).

Textbook has a typo. Sample size should be 17 (instead of 16). Degree of freedom should be 16 (intead of 15).

```{r}
var_test(obs = EXA_C07_S07_01$mass, null.value = 600)
```

## Section 7.8

### Example 7.8.1

Page 268 (10th ed), Page 242 (11th ed)

```{r}
var_test(xsd = c(30.62, 11.37), n = 6L, null.value = 1, alternative = 'greater')
```

### Example 7.8.2

Page 270 (10th ed),

```{r}
EXA_C07_S03_02 |>
  with(expr = var.test(x = CONTROL, y = SCI))
```

## Section 7.9 The Type II Error and the Power of a Test

### Example 7.9.1

Page 272 (10th ed), Page 245 (11th ed). Table 7.9.1

```{r}
seq(from = 16, to = 19, by = .5) |>
  power_z(null.value = 17.5, sd = 3.6, n = 100L)
```

### Example 7.9.2

Page 276 (10th ed), Page 248 (11th ed)

```{r}
seq(from = 50, to = 70, by = 5) |>
   power_z(null.value = 65, sd = 15, n = 20L, sig.level = .01, alternative = 'less')
```

### Exercise 7.9.1

Page 277

```{r}
c(510, 520, 540) |>
  power_z(null.value = 516, sd = 32, n = 16L, alternative = 'greater')
```

### Exercise 7.9.3

Page 277,

```{r}
c(4, 4.5, 4.75, 5) |>
  power_z(null.value = 4.25, sd = 1.8, n = 81L, sig.level = .01, alternative = 'greater')
```

## Section 7.10

### Example 7.10.1

Page 278 (10th ed),

```{r}
(n_d7101 <- uniroot(f = function(x) {
  power_z(55, null.value = 65, sd = 15, n = x, sig.level = .01, alternative = 'less')$power - .95
}, interval = c(0, 50))$root)
power_z(55, null.value = 65, sd = 15, n = ceiling(n_d7101), sig.level = .01, alternative = 'less')
```

### Exercise 7.10.2

Page 279,

```{r}
(n_e7102 = uniroot(f = function(x) {
  power_z(4.52, null.value = 4.5, sd = .02, n = x, sig.level = .01, alternative = 'greater')$power - .95
}, interval = c(0, 50))$root)
```

The required sample size to achieve 95% power at 4.52 is 15.77, or 16 as an integer. The rejection region is (-Inf, 4.51)

```{r}
qnorm(p = .01, mean = 4.5, sd = .02/sqrt(ceiling(n_e7102)), lower.tail = FALSE)
```

## Review

### Review 18

Page 283,

Let group 1 be IVF and group 2 be spontaneous conception

```{r}
t_test(xbar = c(3071, 3172), xsd = c(761, 702), n = c(163L, 321L), 
             null.value = 0, alternative = 'less')
```

### Review 19

Page 283

```{r}
REV_C07_19 |> with(expr = t.test(Before, After, alternative = 'greater', paired = TRUE))
```

### Review 20

Page 284

```{r}
t_test(xbar = c(5.56, 4.40), xsd = c(0.84, 0.57), n = c(40L, 57L), null.value = 0, alternative = 'two.sided')
```

### Review 25

Page 284

```{r}
prop_CLT(x = 27L, n = 90L, null.value = .25, alternative = 'greater')
```

```{r}
prop.test(x = 27L, n = 90L, p = .25, alternative = 'greater')
```

### Review 44

Page 290

```{r}
REV_C07_44 |> head(n = 3L)
```

## Large Data Sets

Page 299,

### Question 1

```{r}
LDS_C07_PCKDATA$A |> shapiro.test()
```

```{r}
LDS_C07_PCKDATA$B |> shapiro.test()
```

```{r}
LDS_C07_PCKDATA |> with(expr = t.test(A, B, paired = TRUE))
```

```{r}
LDS_C07_PCKDATA |> with(expr = t.test(A, B, alternative = 'greater', paired = TRUE))
```

### Question 2

```{r}
boxplot(TIME ~ Group, data = LDS_C07_PROTHROM)
t.test(TIME ~ Group, data = LDS_C07_PROTHROM)
```

### Question 3

```{r}
plot(SCA ~ NC, data = LDS_C07_HEADCIRC)
abline(a = 0, b = 1, col = 'red')
LDS_C07_HEADCIRC |> with(expr = t.test(SCA, NC, alternative = 'less', paired = TRUE))
```

### Question 4

```{r}
boxplot(Hb ~ Group, data = LDS_C07_HEMOGLOB)
t.test(Hb ~ Group, data = LDS_C07_HEMOGLOB)
```

### Question 5

```{r}
boxplot(SCORE ~ Learning, data = LDS_C07_MANDEXT)
t.test(SCORE ~ Learning, data = LDS_C07_MANDEXT, alternative = 'less')
```

# Chapter 8: *Analysis of Variance*

## Section 8.2

### Example 8.2.1

Page 318 (10th ed), Page 280 (11th ed)

```{r}
boxplot(selenium ~ type, data = EXA_C08_S02_01, main = 'Figure 8.2.7')
```

```{r}
aov_d821 = aov(selenium ~ type, data = EXA_C08_S02_01)
aov_d821 |> anova() 
```

### Example 8.2.2

Page 325 (10th ed), Page 286 (11th ed)

```{r}
tukey_d822 = aov_d821 |> TukeyHSD(conf.level = .95)
tukey_d822 # Figure 8.2.8
```

```{r}
tukey_d822 |> plot()
```

## Section 8.3

### Example 8.3.1

Page 339 (10th ed), Page 298 (11th ed)

```{r}
EXA_C08_S03_01 |> 
  within(expr = {
    ageGroup = structure(ageGroup, levels = c('<20', '20s', '30s', '40s', '>50'), class = 'factor')
  }) |> 
  aov(formula = time ~ method + ageGroup) |>
  anova()
```

## Section 8.4

### Example 8.4.1

Page 348 (10th ed), Page 307 (11th ed)

```{r}
EXA_C08_S04_01 |>
  within(expr = {
    SUBJ = factor(SUBJ)
    TIME = structure(TIME, levels = c('Baseline', '1-Mon', '3-Mon', '6-Mon'), class = 'factor')
  }) |> 
  aov(formula = FUNC ~ SUBJ + TIME) |> 
  anova()
```

### Example 8.4.2

Page 352 (10th ed), Page 310 (11th ed).

```{r}
EXA_C08_S04_02 |> head(n = 3L)
```

```{r}
d842 = EXA_C08_S04_02 |> 
  within(expr = {
    subject = factor(subject)
    treatment = structure(treatment, levels = c('placebo', 'aloe_juice'), class = 'factor')
  }) |>
  reshape2::melt(id.vars = c('subject', 'treatment'), variable.name = 'time', value.name = 'OralScores')
d842 |> head(n = 3L)
```

Hypotheses: - Main effect of 'treatment'; - Main effect of 'time'; - Interaction between 'treatment' and 'time'

```{r}
d842 |> 
  aov(formula = OralScores ~ treatment * time + Error(subject)) |>
  summary()
# Section 'Error: subject' in R output
# .. is Figure 8.4.4 'Tests of Between-Subjects Effects' (without the row of 'Intercept')
# .. 'treatment' row: effect of treatment at **baseline** (i.e. reference time), 
# ... degree-of-freedom (dof) = 2-1 = 1
# .. 'Residuals' row: residual at baseline, dof = (25-1) - (2-1) = 23
# .. It's important to note that 'treatment' is a **between-subject factor**.
# Section 'Error: Within' in R output
# .. is Figure 8.4.4 'Tests of Within-Subjects Effects'
# .. 'time' row: effect of time within subject for placebo (i.e. reference treatment), dof = 4-1 = 3
# .. 'treatment:time' row: interation of treatment and time, dof = (2-1)*(4-1) = 3
# .. 'Residuals' row: residual at 2wk, 4wk and 6wk, dof = (4-1)*23 = 69 
# ... [(4-1) timepoints, 23 dof at each timepoints]
# Analysis Interpretation
# .. No signif. diff. detected between placebo vs. aloe at baseline (p = .815)
# .. No signif. diff. detected in the trends over time between placebo vs. aloe (p = .974)
# .. Signif. diff. detected among the four timepoints, for either placebo or aloe pts (p = 3e-7)
```

An equivalent ANOVA model

```{r}
d842 |> 
  aov(formula = OralScores ~ treatment * time + subject) |> 
  anova()
```

## Section 8.5

### Example 8.5.2

Page 364 (10th ed), Page 321 (11th ed)

```{r}
EXA_C08_S05_02 |> 
  within(expr = {
    A = structure(A, levels = c('Cardiac', 'Cancer', 'CVA', 'Tuberculosis'), class = 'factor')
    B = structure(B, levels = c('20s', '30s', '40s', '50+'), class = 'factor')
  }) |> 
  aov(formula = HOME ~ A * B) |>
  anova()
```

## Review

### Review 13

Page 376

```{r}
REV_C08_13 |> head(n = 3L)
```

```{r}
r813 = REV_C08_13 |> 
  within(expr = {
    Group = factor(Group, levels = unique.default(Group))
  })
```

```{r}
boxplot(Count ~ Group, data = r813)
```

```{r}
aov_r813 = aov(Count ~ Group, data = r813)
aov_r813 |> anova()
```

```{r}
aov_r813 |> TukeyHSD()
```

### Review 14

Page 377,

```{r}
REV_C08_14 |> head(n = 3L)
```

```{r}
r814 = REV_C08_14 |> 
  within(expr = {
    GROUP = structure(GROUP, levels = c('Control', 'Sham', 'OMT'), class = 'factor')
  })
```

```{r}
boxplot(SCALE ~ GROUP, data = r814)
```

```{r}
aov_r814 = aov(SCALE ~ GROUP, data = r814)
aov_r814 |> anova()
```

```{r}
aov_r814 |> TukeyHSD()
```

### Review 16

Page 379 (10th ed),

```{r}
REV_C08_16 |> head(n = 3L)
```

```{r}
aov(Count ~ Sens * Treat, data = REV_C08_16) |> 
  anova()
```

```{r}
aov(Count ~ Sens + Treat, data = REV_C08_16) |>
  anova()
```

### Review 17

Page 380,

```{r}
REV_C08_17 |> head(n = 3L)
```

```{r}
REV_C08_17 |> 
  within(expr = {
    SMOKE = structure(SMOKE, levels = c('none', 'light', 'heavy'), class = 'factor')
    VEG = structure(as.integer(gsub('^ve', replacement = '', VEG)), levels = c('fewest', 'medium', 'most'), class = 'factor')
  }) |>
  aov(formula = AMP ~ SMOKE * VEG) |> 
  anova()
```

## Large Data Sets

Page 407-408,

### Question 1

```{r}
MedScore = LDS_C08_MEDSCORES |> 
  within(expr = {
    day = factor(day)
    test = factor(test)
    ID = factor(ID)
  })
aov(score ~ day * test + ID, data = MedScore) |>
  anova()
```

### Question 2

```{r}
LDS_C08_LSADATA |> head(n = 3L)
```

```{r}
LSA1 = LDS_C08_LSADATA |>
  reshape2::melt(id.vars = 'subj', variable.name = 'Group', value.name = 'LSA')
```

```{r}
boxplot(LSA ~ Group, data = LSA1)
```

```{r}
aov_LSA = aov(LSA ~ Group, data = LSA1)
aov_LSA |> anova()
```

```{r}
aov_LSA |> TukeyHSD()
```

### Question 3

```{r}
LDS_C08_SACEDATA |> head(n = 3L)
```

```{r}
SACE = LDS_C08_SACEDATA |>
  reshape2::melt(id.vars = 'OBSERV', variable.name = 'Group', value.name = 'SACE')
```

```{r}
boxplot(SACE ~ Group, data = SACE)
```

```{r}
aov_SACE = aov(SACE ~ Group, data = SACE)
aov_SACE |> anova()
```

```{r}
aov_SACE |> TukeyHSD()
```

### Question 4

```{r}
LDS_C08_CSFDATA |> head(n = 3L)
```

```{r}
CFA = LDS_C08_CSFDATA |>
  reshape2::melt(id.vars = 'OBSERV', variable.name = 'Disease', value.name = 'CSF')
```

```{r}
boxplot(CSF ~ Disease, data = CFA)
```

```{r}
aov_CSF = aov(CSF ~ Disease, data = CFA)
aov_CSF |> anova() 
```

```{r}
aov_CSF |> TukeyHSD()
```

### Question 5

```{r}
LDS_C08_RBCDATA |> head(n = 3L)
```

```{r}
RBCDATA = LDS_C08_RBCDATA |>
  reshape2::melt(id.vars = 'OBSERV', variable.name = 'Diet', value.name = 'RBC')
```

```{r}
boxplot(RBC ~ Diet, data = RBCDATA)
```

```{r}
aov_RBC = aov(RBC ~ Diet, data = RBCDATA)
aov_RBC |> anova() 
```

```{r}
aov_RBC |> TukeyHSD()
```

### Question 6

```{r}
LDS_C08_SERUMCHO |> head(n = 3L)
```

```{r}
SERUMCHO1 = LDS_C08_SERUMCHO |>
  within(expr = {
    SUBJ = factor(SUBJ)
  })
```

```{r}
SERUMCHO2 = SERUMCHO1 |>
  reshape2::melt(id.var = 'SUBJ', variable.name = 'Diet', value.name = 'SC')
```

```{r}
aov_SERUMCHO = aov(SC ~ Diet + SUBJ, data = SERUMCHO2) # 'SUBJ' as block factor
```

```{r}
aov_SERUMCHO |> anova()
```

# Chapter 9: *Simple Linear Regression and Correlation*

## Section 9.3

### Example 9.3.1

Page 417 (10th ed), Page 358 (11th ed)

```{r}
EXA_C09_S03_01 |> head(n = 3L)
```

```{r}
m931 = lm(AT ~ Waist, data = EXA_C09_S03_01)
plot(AT ~ Waist, data = EXA_C09_S03_01, 
     xlab = 'Waist circumference (cm), X', 
     ylab = 'Deep abdominal AT area (cm2), Y', 
     main = 'Figure 9.3.1 and 9.3.3'); abline(m931, col = 'red', lty = 2)
```

## Section 9.4

### Example 9.4.1

Page 432 (10th ed), Page 372 (11th ed) Table 9.4.1. Page 434 (10th ed).

```{r}
m931 |> anova()
```

### Example 9.4.2

Page 436 (10th ed), Page 375 (11th ed)

```{r}
m931 |> summary()
```

Page 438 (10th ed).

```{r}
m931 |> confint()
```

### Example 9.4.3

Page 376 (11th ed). Omitted

### Example 9.4.3; 9.4.4

Page 440 (10th ed). Page 379 (11th ed)

```{r}
m931 |> plot(which = 1, main = 'Figure 9.4.8 (10th) or 9.4.9 (11th)')
```

## Section 9.7

### Example 9.7.1

Page 447 (10th ed), Page 386 (11th ed)

```{r}
EXA_C09_S07_01 |> head(n = 3L)
```

```{r}
plot(CV ~ HEIGHT, data = EXA_C09_S07_01, xlab = 'Height (cm)', ylab = 'Cv (units)')
```

```{r}
m971 = lm(CV ~ HEIGHT, data = EXA_C09_S07_01)
```

```{r}
m971 |> summary()
```

Figure 9.7.3; Page 450 (10th ed)

```{r}
m971 |> anova()
```

### Example 9.7.2

Page 452 (10th ed), Page 390 (11th ed)

```{r}
EXA_C09_S07_01 |> cor()
```

Figure 9.7.4, 9.7.5.

```{r}
cor.test(~ CV + HEIGHT, data = EXA_C09_S07_01)
```

Page 453, when the hypothesized rho ss a nonzero value; R does not have a function to do this.

library(DanielBiostatistics10th) library(reshape2)

library(DanielBiostatistics10th)

## Review

### Review 17

Page 465,

```{r}
REV_C09_17 |> head(n = 3L)
```

```{r}
plot(BOARD ~ AVG, data = REV_C09_17)
```

```{r}
lm(BOARD ~ I(AVG - 85), data = REV_C09_17) |> 
  summary() # ideally
```

```{r}
lm2_917 = lm(BOARD ~ AVG, data = REV_C09_17)  # also acceptable
lm2_917 |> summary()
```

```{r}
plot(BOARD ~ AVG, data = REV_C09_17); abline(reg = lm2_917)
```

### Review 18

Page 466,

```{r}
REV_C09_18 |> head(n = 3L)
```

```{r}
lm(ATT ~ AGE, data = REV_C09_18) |> summary()
```

```{r}
plot(ATT ~ AGE, data = REV_C09_18)
```

```{r}
cor.test(~ ATT + AGE, data = REV_C09_18)
```

```{r}
lm(SOC ~ AGE, data = REV_C09_18) |> 
  summary()
```

```{r}
plot(SOC ~ AGE, data = REV_C09_18)
```

```{r}
cor.test(~ SOC + AGE, data = REV_C09_18)
```

```{r}
lm(HYP ~ AGE, data = REV_C09_18) |> summary()
```

```{r}
plot(HYP ~ AGE, data = REV_C09_18)
```

```{r}
cor.test(~ HYP + AGE, data = REV_C09_18)
```

### Review 19

```{r}
lm(TIME ~ PROBLEMS, data = REV_C09_19) |> summary()
```

### Review 29

Page 471,

```{r}
REV_C09_29 |> head(n = 3L)
```

Use all 85 pairs of measurements (textbook asks you to select the subset of B\<1000; ignore)

```{r}
# r929b <- subset(REV_C09_29, Standard < 1000)
```

```{r}
m_r929 = lm(New ~ Standard, data = REV_C09_29)
m_r929 |> summary()
```

```{r}
m_r929 |> confint()
```

```{r}
plot(New ~ Standard, data = REV_C09_29); abline(m_r929)
```

## Large Data Sets

Page 485-486,

### Question 1

```{r}
LDS_C09_CEREBRAL |> head(n = 3L)
```

```{r}
plot(Pressure ~ Glycerol, data = LDS_C09_CEREBRAL)
```

```{r}
cor.test(~ Pressure + Glycerol, data = LDS_C09_CEREBRAL)
```

```{r}
CEREBRAL = lm(Pressure ~ Glycerol, data = LDS_C09_CEREBRAL)
CEREBRAL |> summary()
```

```{r}
CEREBRAL |> anova()
```

```{r}
plot(Pressure ~ Glycerol, data = LDS_C09_CEREBRAL); abline(reg = CEREBRAL, col = 'red')
```

```{r}
lm(Pressure ~ I(Glycerol - 15), data = LDS_C09_CEREBRAL) |> 
  summary()
```

### Question 2

```{r}
LDS_C09_HYPERTEN |> head(n = 3L)
```

```{r}
HYPERTEN = lm(sbp ~ dbp, data = LDS_C09_HYPERTEN)
HYPERTEN |> summary()
```

```{r}
plot(sbp ~ dbp, data = LDS_C09_HYPERTEN)
abline(reg = HYPERTEN, col = 'red')
```

### Question 3

```{r}
LDS_C09_CALCIUM |> head(n = 3L)
LDS_C09_CALCIUM_a = LDS_C09_CALCIUM |>
  reshape2::melt(id.vars = 'patient', variable.name = 'Prednisolone', value.name = 'BodyCalcium')
```

```{r}
boxplot(BodyCalcium ~ Prednisolone, data = LDS_C09_CALCIUM_a)
```

```{r}
CALCIUM = lm(BodyCalcium ~ Prednisolone, data = LDS_C09_CALCIUM_a)
CALCIUM |> summary()
```

# Chapter 10: *Multiple Regression and Correlation*

## Section 10.3

### Example 10.3.1

Page 493 (10th ed), Page 419 (11th ed)

Page 502 (10th ed), Page 428 (11th ed)

Page 504 (10th ed), Page 429 (11th ed)

Page 505 (10th ed), Page 430 (11th ed)

Page 506 (10th ed), Page 431 (11th ed)

```{r}
EXA_C10_S03_01 |> pairs(main = 'Figure 10.3.1')
```

see 'Multiple R-squared' (not 'Adjusted R-squared')

see 'F-statistic'

see 'Coefficients:'

```{r}
m1031 = lm(CDA ~ AGE + EDLEVEL, data = EXA_C10_S03_01)
```

```{r}
m1031 |> summary()
```

confidence interval for beta's;

```{r}
m1031 |> confint()
```

## Section 10.4

### Example 10.4.1-10.4.3

## Section 10.5

### Example 10.5.1

Page 509 (10th ed), Page 434 (11th ed)

```{r}
(newd_1031 = data.frame(AGE = 68, EDLEVEL = 12))
```

```{r}
m1031 |> predict(newdata = newd_1031, interval = 'prediction')
```

```{r}
m1031 |> predict(newdata = newd_1031, interval = 'confidence')
```

## Section 10.6

### Example 10.6.1

Page 511 (10th ed), Page 436 (11th ed)

```{r}
EXA_C10_S06_01 |> pairs()
m1061 = lm(W ~ P + S, data = EXA_C10_S06_01)
```

```{r}
m1061 |> summary()
```

```{r}
m1061 |> confint()
```

```{r}
m1061 |> anova()
```

### Example 10.6.2;

Page 515 (10th ed), Page 440 (11th ed)

```{r}
psych::partial.r(EXA_C10_S06_01, x = 2:3, y = 1L)
```

library(DanielBiostatistics10th) library(car)

## Large Data Sets

Page 536,

### Question 1

```{r}
LDS_C10_LTEXER |> head(n = 3L)
```

### Question 2

```{r}
LDS_C10_STERLENGTH |> head(n = 3L)
```

### Question 3

```{r}
LDS_C10_RISKFACT |> head(n = 3L)
```

```{r}
car::scatterplotMatrix(~ O2 + SBP + TotalChol + HDLChol + TriG, data = LDS_C10_RISKFACT, smooth = FALSE, regLine = FALSE)
# There are obvious typos in SBP, the corresponding records should be removed
```

```{r}
LDS_C10_RISKFACT_a = LDS_C10_RISKFACT[-c(
  which.max(LDS_C10_RISKFACT$SBP)
), , drop = FALSE]
```

```{r}
car::scatterplotMatrix(~ O2 + SBP + TotalChol + HDLChol + TriG, data = LDS_C10_RISKFACT_a, smooth = FALSE, regLine = FALSE)
```

```{r}
lm(O2 ~ SBP + TotalChol + HDLChol + TriG, data = LDS_C10_RISKFACT_a) |> 
  summary()
```

### Question 4

This is a wong data, omitted in 11th edition.

```{r}
LDS_C10_RESPDIS |> head(n = 3L)
```

# Chapter 11: *Regression Analysis: Some Additional Techniques*

## Section 11.1

### Example 11.1.1

Page 540,

```{r}
EXA_C11_S01_01 |> head(n = 3L)
```

```{r}
EXA_C11_S01_01$conc |> log(base = 10) |> head(n = 3L)
```

### Example 11.1.2

Page 542,

```{r}
cor.test(~ sbp + weight, data = EXA_C11_S01_02)
```

```{r}
cor.test(~ sbp + bmi, data = EXA_C11_S01_02) 
```

## Section 11.2

### Example 11.2.1

Page 545,

```{r}
d1121 = EXA_C11_S02_01 |>
  within(expr = {
    SMOKE = ifelse(SMOKE, yes = 'Yes', no = 'No')
  })
```

```{r}
car::scatterplot(GRAMS ~ WEEKS | SMOKE, data = d1121, smooth = FALSE,
                 xlab = 'Length of gestation (weeks)', ylab = 'Birth weight (grams)')
```

```{r}
m1121 = lm(GRAMS ~ WEEKS + SMOKE, data = d1121)
```

```{r}
m1121 |> summary()
```

```{r}
m1121 |> confint()
```

### Example 11.2.3

Page 551,

```{r}
d1123 = EXA_C11_S02_03 |> 
  within(expr = {
    METHOD = factor(METHOD, levels = c('C', 'A', 'B')) # textbook designated 'C' as reference level
  })
m1123 = lm(EFFECT ~ AGE * METHOD, data = d1123)
m1123 |> summary() # Figure 11.2.5
m1123 |> confint()
car::scatterplot(EFFECT ~ AGE | METHOD, data = d1123, smooth = FALSE,
                 xlab = 'Age', ylab = 'Treatment effectiveness')
```

### Exercise 11.2.2

Page 557

```{r}
EXR_C11_S02_02 |> head(n = 3L)
```

```{r}
e1122 = EXR_C11_S02_02 |>
  within(expr = {
    DISEASE = structure(DISEASE, levels = c('Cholelithiasis', 'GallbladderCancer'), class = 'factor')
  })
```

```{r}
car::scatterplot(MDA ~ CYTO | DISEASE, data = e1122, smooth = FALSE,
                 xlab = 'Cytochrome P-450 (nmol/mg)', ylab = 'Melonaldehyde (\u03bcg/mg)')
```

```{r}
lm(MDA ~ CYTO * DISEASE, data = e1122) |> summary() 
```

```{r}
lm(MDA ~ CYTO + DISEASE, data = e1122) |> summary()
```

```{r}
lm(MDA ~ DISEASE, data = e1122) |> summary()
```

## Section 11.3

### Example 11.3.1

Page 561,

```{r}
EXA_C11_S03_01 |> head(n = 3L)
```

```{r}
m1131_raw = lm(JOBPER ~ ASRV + ENTH + AMB + COMM + PROB + INIT, data = EXA_C11_S03_01)
m1131_raw |> summary()
# summary(m1131 <- MASS::stepAIC(m1131_raw, direction = 'backward'))
# the stepwise selection criterion used in MINITAB is not necessarily AIC
```

## Section 11.4 Logistic Regression

### Example 11.4.1

Page 572,

```{r}
d1141 = c(92L, 21L, 15L, 20L) |> 
  matrix(nrow = 2L, dimnames = list(
    OCAD = c('Present', 'Absent'), 
    Sex = c('Male', 'Female')
  )) |>
  as.table()
```

```{r}
d1141 |> as_flextable()
```

```{r}
(d1141a = d1141 |> 
  as.data.frame.table() |> 
  within(expr = {
    OCAD = (OCAD == 'Present')
    Sex = factor(Sex, levels = c('Female', 'Male'))
  }))
```

```{r}
m1141 = glm(OCAD ~ Sex, family = binomial, weights = Freq, data = d1141a)
```

```{r}
m1141 |> summary()
```

```{r}
m1141$coefficients[2L] |> exp() # exp(beta_M)
```

Confidence interval of $\exp(\mathbf{\beta})$

```{r}
#| message: false
m1141 |> confint() |> exp() 
```

```{r}
m1141 |> predict(newdata = data.frame(Sex = setNames(nm = c('Male', 'Female'))), type = 'response')
```

### Example 11.4.2

Page 573,

```{r}
EXA_C11_S04_02 |> head(n = 3L)
```

```{r}
summary(m1142 <- glm(ATT ~ AGE, family = binomial, data = EXA_C11_S04_02))
```

```{r}
m1142$coefficients[2L] |> exp()
```

```{r}
#| message: false
m1142 |> confint() |> exp()
```

```{r}
m1142 |> car::Anova() # Optional
```

### Example 11.4.3

Page 576,

```{r}
REV_C11_24 |> head(n = 3L)
```

```{r}
glm(ONSET ~ HIAA + TRYPT, family = binomial, data = REV_C11_24) |> 
  summary() # Figure 11.4.4
# Predictor TRYPT should be removed from model due to p-value \approx 1 
```

```{r}
glm(ONSET ~ HIAA, family = binomial, data = REV_C11_24) |> 
  summary() 
```

### Example 11.4.4-11.4.5

Page 578-579

```{r}
m1142 |> DescTools::PseudoR2(which = c('CoxSnell', 'Nagelkerke'))
```

### Exercise 11.4.1

Page 581

```{r}
e1141 = c(611L, 68L, 308L, 37L) |> 
  matrix(nrow = 2L, dimnames = list(
    Victimizated = c('No', 'Yes'), 
    Sex = c('Women', 'Man')
  )) |> 
  as.table()
```

```{r}
e1141 |> as_flextable()
```

```{r}
e1141 |> 
  as.data.frame.table() |> 
  within(expr = {
    Victimizated = (Victimizated == 'Yes')
  }) |> 
  glm(formula = Victimizated ~ Sex, family = binomial, weights = Freq) |> 
  summary()
```

### Exercise 11.4.2

Page 581,

```{r}
EXR_C11_S04_02 |> head(n = 3L)
```

```{r}
e1142 = glm(PART ~ INDEX, family = binomial, data = EXR_C11_S04_02)
e1142 |> summary()
```

```{r}
e1142$coefficients |> exp()
```

```{r}
e1142 |> confint() |> exp()
```

## Review

### Review 15

Page 585 (10th ed)

```{r}
REV_C11_15 |> head(n = 3L)
```

```{r}
lm(CARDIAC ~ VO2 * GROUP, data = REV_C11_15) |> summary()
```

```{r}
m1115 = lm(CARDIAC ~ VO2 + GROUP, data = REV_C11_15)
m1115 |> summary()
```

```{r}
m1115 |> confint()
```

```{r}
car::scatterplot(CARDIAC ~ VO2 | GROUP, data = REV_C11_15, smooth = FALSE,
                 xlab = 'Oxygen Consumption (VO2)', ylab = 'Cardiac Output (L/min)')
```

### Review 16

Page 585 (10th ed)

```{r}
REV_C11_16 |> head(n = 3L)
```

```{r}
m1116 = lm(POTAS ~ WATER * SEX, data = REV_C11_16)
m1116 |> summary()
```

```{r}
car::scatterplot(POTAS ~ WATER | SEX, data = REV_C11_16, smooth = FALSE,
                 xlab = 'Total Body Water (L)', ylab = 'Total Body Potassium (mEq)')
```

### Review 17

Page 586

```{r}
REV_C11_17 |> head(n = 3L)
```

```{r}
car::scatterplot(WEIGHT ~ WEEKS | HOSP, data = REV_C11_17, smooth = FALSE)
```

```{r}
lm(WEIGHT ~ WEEKS * HOSP, dat = REV_C11_17) |> 
  summary()
```

### Review 23

Page 589,

```{r}
REV_C11_23 |> head(n = 3L)
```

```{r}
car::scatterplot(V ~ W | GROUP, data = REV_C11_23, smooth = FALSE)
```

```{r}
lm(V ~ W * GROUP, data = REV_C11_23) |>
  summary()
```

### Review 24

Page 590.

see Example 11.4.3

## Large Data Sets

Page 596-597,

### Question 1

```{r}
LDS_C11_AQUATICS |> head(n = 3L)
```

### Question 2

```{r}
LDS_C11_TEACHERS |> head(n = 3L)
```

### Question 3

```{r}
LDS_C11_WGTLOSS |> head(n = 3L)
```

```{r}
LDS_C11_WGTLOSS$Group |> table()
```

```{r}
car::scatterplot(Protein ~ IdealBW | Group, data = LDS_C11_WGTLOSS, smooth = FALSE)
```

```{r}
lm(Protein ~ IdealBW * Group, data = LDS_C11_WGTLOSS) |>
  summary()
```

```{r}
lm(Protein ~ IdealBW + Group, data = LDS_C11_WGTLOSS) |> 
  summary()
```

# Chapter 12: The $\chi^2$-Distribution and the Analysis of Frequencies

## Section 12.3

### Example 12.3.1

Page 605 (10th ed)

```{r}
d1231_b = c(-Inf, seq(from = 125, to = 275, by = 25), Inf)
d1231 = c(1L, 3L, 8L, 18L, 6L, 4L, 4L, 3L)
names(d1231) = cut(double(), breaks = d1231_b, right = FALSE, include.lowest = TRUE) |> levels()
```

```{r}
p1231 = d1231_b |> 
  pnorm(mean = 198.67, sd = 41.31) |> 
  diff()
p1231 |> 
  label_percent(accuracy = .1)()
```

```{r}
#| warning: false
(m = d1231 |> 
   chisq.test(p = p1231) |>
   update_df(restriction = c('estimating mean', 'estimating sd')))
```

```{r}
m |> viewOE()
```

### Example 12.3.2

Page 609 (10th ed),

```{r}
# 100 doctors, 25 patients per doctor
d1232 = c(5L, 6L, 8L, 10L, 10L, 15L, 17L, 10L, 10L, 9L, 0L)
o1232 = setNames(c(sum(d1232[1:2]), d1232[-(1:2)]), nm = c('0-1', 2:9, '10 or more'))
```

```{r}
(p = sum((0:10) * d1232) / (25 * 100)) # binomial `prob`
p1232 = c(
  pbinom(1L, size = 25L, prob = p),
  dbinom(2:9, size = 25L, prob = p),
  pbinom(9, size = 25L, prob = p, lower.tail = FALSE))
p1232 |>
  label_percent(accuracy = .1)()
```

```{r}
#| warning: false
(m1 = o1232 |> 
  chisq.test(p = p1232) |>
  update_df(restriction = 'estimating `p`'))
```

```{r}
m1 |> viewOE()
```

### Example 12.3.3

Page 611 (10th ed),

```{r}
d1233 = c(5L, 14L, 15L, 23L, 16L, 9L, 3L, 3L, 1L, 1L, 0L)
o_1233 = setNames(c(d1233[1:8], sum(d1233[-(1:8)])), nm = c(0:7, '8 or more'))
```

```{r}
p1233 = c(
  dpois(0:7, lambda = 3), # lambda = 3 is provided by the textbook
  ppois(7L, lambda = 3, lower.tail = FALSE))
p1233 |>
  label_percent(accuracy = .1)()
```

```{r}
#| warning: false
(m = o_1233 |>
  chisq.test(p = p1233))
```

```{r}
m |> viewOE()
```

### Example 12.3.4

Page 614 (10th ed), Page 531 (11th ed)

```{r}
d1234 = c('Dec 05' = 62L, 'Jan 06' = 84L, 'Feb 06' = 17L, 'Mar 06' = 16L, 'Apr 06' = 21L)
```

```{r}
(m = d1234 |> chisq.test())
```

```{r}
m |> viewOE()
```

### Example 12.3.5

Page 616 (10th ed), Page 533 (11th ed)

```{r}
(m = c(dominant = 43L, heterozygous = 125L, recessive = 32L) |> 
   chisq.test(p = c(1, 2, 1), rescale.p = TRUE))
```

```{r}
m |> viewOE()
```

## Section 12.4

### Example 12.4.1

Page 621 (10th ed)

```{r}
d1241 = c(260L, 15L, 7L, 299L, 41L, 14L) |>
  matrix(nrow = 3L, dimnames = list(
    Race = c('White', 'Black', 'Other'),
    FolicAcid = c('TRUE', 'FALSE')))
```

```{r}
d1241 |> as.table() |> as_flextable()
```

```{r}
d1241 |> chisq.test()
```

### Example 12.4.2

Page 626 (10th ed),

```{r}
d1242 = c(131L, 14L, 52L, 36L) |>
  matrix(nrow = 2L, dimnames = list(
    Type = c('Faller', 'Non-Faller'),
    LifestyleChange = c('TRUE', 'FALSE')))
```

```{r}
d1242 |> as.table() |> as_flextable()
```

```{r}
d1242 |> chisq.test(correct = FALSE)
```

```{r}
d1242 |> chisq.test(correct = TRUE)
```

### Exercise 12.4.4

```{r}
c(17,25,96,97,142,816) |> 
  array(dim = c(3L, 2L), dimnames = list(
    Weight = c('Underweight', 'Overweight', 'Appropriate'),
    Smoking = c('Yes', 'No')
  )) |>
  as.table() |>
  chisq.test()
```

## Section 12.5

### Example 12.5.1

Page 631 (10th ed),

```{r}
d1251 = c(21L, 19L, 75L, 77L) |>
  matrix(nrow = 2L, dimnames = list(
    Group = c('Narcoleptic', 'Healthy'),
    Migraine = c('TRUE', 'FALSE')))
```

```{r}
d1251 |> as.table() |> as_flextable()
```

```{r}
d1251 |> chisq.test(correct = FALSE)
```

## Section 12.6

### Example 12.6.1

Page 638 (10th ed),

```{r}
d1262 = c(2L, 8L, 7L, 4L) |>
  matrix(nrow = 2L, dimnames = list(
    Group = c('PI Naive', 'PA Experienced'),
    Regimen2yr = c('TRUE', 'FALSE')))
```

```{r}
d1262 |> as.table() |> as_flextable()
```

```{r}
d1262 |> fisher.test()
```

## Section 12.7

### Example 12.7.1

Page 644 (10th ed),

```{r}
d1271 = c(22L, 18L, 216L, 199L) |>
  matrix(nrow = 2L, dimnames = list(
    Exercising = c('Extreme', 'No'),
    PretermLabor = c('Cases', 'Noncases')))
```

```{r}
d1271 |> as.table() |> as_flextable()
```

```{r}
d1271 |> DescTools::RelRisk(conf.level = .95)
```

```{r}
d1271 |> fisher.test()
```

### Example 12.7.2

Page 647 (10th ed),

```{r}
d1272 = c(64L, 68L, 342L, 3496L) |>
  matrix(nrow = 2L, dimnames = list(
    SmkPregnancy = c('Smoked Throughout', 'Never Smoked'),
    Obesity = c('Cases', 'Noncases')))
```

```{r}
d1272 |> as.table() |> as_flextable()
```

```{r}
d1272 |> fisher.test()
```

### Example 12.7.3-12.7.4

Page 650-652 (10th ed),

```{r}
d1273 = c(21L, 16L, 11L, 6L, 50L, 18L, 14L, 6L) |> 
  array(dim = c(2L, 2L, 2L), dimnames = list(
    HTN = c('Present', 'Absent'), 
    OCAD = c('Cases', 'Controls'), 
    Age = c('<=55', '>55')))
```

```{r}
d1273 |> addmargins(margin = 1:2)
```

```{r}
d1273 |> mantelhaen.test()
```

## Review

### Review 17

Page 659,

```{r}
r1217 = c(72L, 54L, 16L, 8L, 230L, 192L, 63L, 15L) |>
  matrix(nrow = 4L, dimnames = list(
    BloodGroup = c('O', 'A', 'B', 'AB'), 
    Antigen = c('Carriers', 'Noncarriers'))) 
```

```{r}
r1217 |> as.table() |> as_flextable()
```

```{r}
r1217 |> chisq.test()
```

### Review 37

Page 664,

```{r}
r1237 = c(29L, 9L, 123L-29L, 192L-123L-9L) |>
  matrix(nrow = 2L, dimnames = list(Gender = c('Male', 'Female'), Drug = c('Yes', 'No')))
```

```{r}
r1237 |> as.table() |> as_flextable()
```

```{r}
r1237 |> chisq.test()
```

## Large Data Sets

Page 665-666

```{r}
LDS_C12_SMOKING |> head(n = 3L)
```

```{r}
SMOKING1 = LDS_C12_SMOKING |>
  within(expr = {
    Gender = structure(Gender + 1L, levels = c('Male', 'Female'), class = 'factor')
    Smoking = ifelse(Smoking, yes = 'yes', no = 'no')
    Drinking = structure(Drinking + 1L, levels = c('None', 'Light', 'Heavy'), class = 'factor')
    Respiratory = ifelse(Respiratory, yes = 'yes', no = 'no')
    HighBloodPres = ifelse(HighBloodPres, yes = 'yes', no = 'no')
  })
```

### Question 1

```{r}
SMOKING1 |> with(expr = chisq.test(x = Smoking, y = Respiratory))
```

### Question 2

```{r}
SMOKING1 |> with(expr = chisq.test(x = Drinking, y = HighBloodPres))
```

### Question 3

```{r}
SMOKING1 |> with(expr = chisq.test(x = Smoking, y = Gender))
```

### Question 4

```{r}
SMOKING1 |> with(expr = chisq.test(x = Drinking, y = Gender))
```

# Chapter 13: *Nonparametric and Distribution-Free Statistics*

## Section 13.3 *The Sign Test*

### Example 13.3.1

Page 673

```{r}
c(4, 5, 8, 8, 9, 6, 10, 7, 6, 6) |>
  BSDA::SIGN.test(md = 5)
```

### Example 13.3.2

Page 677,

```{r}
EXA_C13_S03_02 |> 
  with(expr = {
    BSDA::SIGN.test(x = X, y = Y, alternative = 'less')
  })
```

## Section 13.4 *The Wilcoxon Signed-Rank Test for Location*

### Example 13.4.1

Page 683

```{r}
c(4.91, 4.10, 6.74, 7.27, 7.42, 7.50, 6.56, 4.64, 5.98, 3.14, 3.23, 5.80, 6.17, 5.39, 5.77) |>
  wilcox.test(mu = 5.05)
```

## Section 13.5 *The Median Test*

### Example 13.5.1

Page 686,

```{r}
(med1351 = EXA_C13_S05_01 |> unlist() |> median(na.rm = TRUE)) # common median
t1351 = EXA_C13_S05_01 |> 
  with(expr = {
    tmp = cbind(Urban = table(URBAN < med1351), Rural = table(RURAL < med1351, useNA = 'no'))
    rownames(tmp) = paste('Number of scores', c('above', 'below'), 'median')
    as.table(tmp)
  }) 
```

```{r}
t1351 |> as_flextable()
```

```{r}
t1351 |> chisq.test(correct = FALSE)
```

## Section 13.6 *The Mann-Whitney Test*

### Example 13.6.1;

Page 691,

```{r}
EXA_C13_S06_01 |> 
  with(expr = wilcox.test(X, Y, exact = FALSE, alternative = 'less'))
```

## Section 13.7

### Example 13.7.1;

Page 699,

```{r}
EXA_C13_S07_01$GLUCOSE |>
  ks.test(y = pnorm, mean = 80, sd = 6)
```

## Section 13.8

### Example 13.8.1;

Page 705 (10th ed), Page 611 (11th ed)

```{r}
data.frame(Air = c(12.22, 28.44, 28.13, 38.69, 54.91), 
           Benzaldehyde = c(3.68, 4.05, 6.47, 21.12, 3.33), 
           Acetaldehyde = c(54.36, 27.87, 66.81, 46.27, 30.19)) |>
  reshape2::melt(
    id.vars = NULL, 
    value.name = 'Eosinophil', 
    variable.name = 'Exposure'
  ) |>
  with(expr = kruskal.test(x = Eosinophil, g = Exposure))
```

### Example 13.8.2;

Page 708 (10th ed), Page 613 (11th ed)

```{r}
EXA_C13_S08_02 |>
  reshape2::melt(
    id.vars = NULL, 
    value.name = 'Value', 
    variable.name = 'Hospital'
  ) |>
  with(expr = kruskal.test(x = Value, g = Hospital))
```

## Section 13.9

### Example 13.9.1;

Page 713 (10th ed); Page 618 (11th ed)

```{r}
EXA_C13_S09_01[LETTERS[1:3]] |>
  as.matrix(rownames.force = TRUE) |>
  friedman.test()
```

### Example 13.9.2;

Page 715 (10th ed); Page 620 (11th ed)

```{r}
EXA_C13_S09_02[LETTERS[1:4]] |>
  as.matrix(rownames.force = TRUE) |>
  friedman.test()
```

## Section 13.10

### Example 13.10.1;

Page 720,

```{r}
EXA_C13_S10_01 |> 
  with(expr = cor.test(X, Y, method = 'spearman'))
```

### Example 13.10.2;

Page 722,

```{r}
EXA_C13_S10_02 |> 
  with(expr = cor.test(V2, V3, method = 'spearman', exact = FALSE))
```

## Section 13.11

### Example 13.11.1-13.11.2;

Page 728-729,

```{r}
testosterone = c(230, 175, 315, 290, 275, 150, 360, 425)
citricAcid = c(421, 278, 618, 482, 465, 105, 550, 750)
```

```{r}
robslopes::TheilSen(x = citricAcid, y = testosterone, verbose = FALSE)
# textbook uses *central* median of ordered pairs, while ?robslopes::TheilSen uses *upper* median
```

```{r}
mblm::mblm(testosterone ~ citricAcid, repeated = FALSE) |> 
  summary() # Figure 13.11.1 (11th ed)
```

## Large Data Sets

Page 747

```{r}
LDS_C13_KLETTER |> head(n = 3L)
```

# Chapter 14: *Survival Analysis*

## Section 14.3 *The Kaplan–Meier Procedure*

### Example 14.3.1

Page 756 (10th ed), Page 652 (11th ed).

```{r}
d1431 = EXA_C14_S03_01 |> 
  within(expr = {
    TIME = .difftime(TIME, units = 'months')
    OS = Surv(time = TIME, event = (VITAL != 'ned'))
  })
```

```{r}
class(d1431$OS)
```

```{r}
head(d1431$OS)
```

Table 14.3.2

```{r}
sf_1431 = survfit(OS ~ TUMOR, data = d1431)
```

```{r}
sf_1431 |> summary()
```

Figure 14.3.1

```{r}
sf_1431 |> 
  plot(lty = 2:3, xlab = 'month', ylab = 'OS', main = 'Figure 14.3.1')
legend(x = 250, y = .9, legend = c('High', 'Low'), lty = 2:3)
```

Fancy plot via `?ggfortify:::autoplot.survfit`

```{r}
#| warning: false
sf_1431 |> 
  autoplot() + xlim(0, 120) + labs(title = 'Figure 14.3.1')
```

Fancy plot via `?survminer::ggsurvplot`.

```{r}
sf_1431 |> 
  survminer::ggsurvplot(title = 'Figure 14.3.1')
```

## Section 14.4 *Comparing Survival Curves*

### Example 14.4.1

Page 764 (10th ed), Page 659 (11th ed)

```{r}
survdiff(OS ~ TUMOR, data = d1431)
```

## Section 14.5 *Cox Regression: The Proportional Hazards Model*

### Example 14.5.1

Page 769 (10th ed), Page 663 (11th ed)

```{r}
d1451 = EXA_C14_S05_01 |> 
  within(expr = {
    time = .difftime(time, units = 'weeks')
    PFS = Surv(time, status)
    drug = drug |>
      structure(levels = c('Opiate', 'Other'), class = 'factor') |>
      relevel(ref = 'Other')
})
```

```{r}
model1_1451 = coxph(PFS ~ drug + age, data = d1451)
model1_1451 |> summary()
```

'Opiate' has higher hazard compared to Drug='Other' (hazard ratio (HR) = 9.407, $p < .001$) With 'proportional hazard' assumption, we don't need to discuss Opiate vs. Other at any specific time point

```{r}
model1_1451 |> confint()
```

'age' is not significant (p = .772), so it should be removed from the model

```{r}
model2_1451 = coxph(PFS ~ drug, data = d1451)
model2_1451 |> summary()
model2_1451 |> confint()
```

```{r}
sf_1451 = survfit(PFS ~ drug, data = d1451)
sf_1451 |> plot(lty = 2:3, xlab = 'weeks', ylab = 'PFS', main = 'Figure 14.5.1')
legend(x = 35, y = .9, legend = c('Other', 'Opiate'), lty = 2:3)
```

fancy plot

```{r}
sf_1451 |> autoplot() + labs(title = 'Figure 14.5.1')
```

another fancy plot

```{r}
sf_1451 |> 
  survminer::ggsurvplot(title = 'Figure 14.5.1')
```

## Review

### Review 11

Page 774,

```{r}
REV_C14_11 |> head(n = 3L)
```

```{r}
r1411 = REV_C14_11 |> 
  subset(subset = (GRADE != '0')) |> # no death for `Grade == 0`
  within(expr = {
    OS = Surv(DAYS, OUTCOME); DAYS = OUTCOME = NULL
    GRADE = factor(GRADE)
  })
```

```{r}
m1411 = coxph(OS ~ GRADE, data = r1411)
```

```{r}
m1411 |> summary()
```

```{r}
survfit(OS ~ GRADE, data = r1411) |> autoplot()
```

### Review 12

Page 775,

```{r}
REV_C14_12 |> head(n = 3L)
```

```{r}
r1412 = REV_C14_12 |> 
  within(expr = {
    OS = Surv(MONTHS, STATUS == 'D'); MONTHS = STATUS = NULL
    NUMBER = factor(NUMBER)
  })
m1412 = coxph(OS ~ NUMBER, data = r1412)
```

```{r}
m1412 |> summary()
```

```{r}
survfit(OS ~ NUMBER, data = r1412) |> autoplot()
```
