---
title: "Wayne W. Daniel\\'s Biostatistics"
author: Tingting Zhan
format: 
  html:
    page-layout: full
    html-math-method: katex
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: visual
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette of package **`DanielBiostatistics10th`** ([CRAN](https://cran.r-project.org/package=DanielBiostatistics10th), [Github](https://github.com/tingtingzhan/DanielBiostatistics10th)) documents the examples of ***Biostatistics: A Foundation for Analysis in the Health Sciences***, [11th Edition](https://www.wiley.com/en-us/Biostatistics%3A+A+Foundation+for+Analysis+in+the+Health+Sciences%2C+11th+Edition-p-9781119496571) by Wayne W. Daniel and Chad L. Cross.

## Note to Users

Examples in this vignette require that the `search` path has

```{r}
#| message: false
library(DanielBiostatistics10th)
library(scales)
library(survival)
library(ggplot2)
library(ggfortify)
library(survminer)
```

```{r}
#| echo: false
library(knitr) # for tables in this vignette
#options(mc.cores = 1L) # for CRAN submission
```

## Terms and Abbreviations

```{r}
#| echo: false
#| results: asis
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`$`', 'Extract parts of an object', '`` ?base::`$` ``',
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', '<https://cran.r-project.org>',
  '`class`', 'Object class', '`?base::class`',
  '`cut`', 'Convert numeric to factor', '`?base::cut`; `?base::cut.default`',
  '`data.frame`', 'Data frame', '`?base::data.frame`',
  '`dim`', 'Dimensions', '`?base::dim`',
  '`dbinom`, `pbinom`, etc.', 'Binomial density and distribution', '`?stats::dbinom`',
  '`factor`', 'Factor', '`?base::factor`',
  '`flextable`', 'Flexible tables', '`?flextable::flextable`',
  '`head`, `tail`', 'First or last parts of an object', '`?utils::head`',
  '`kurtosis`', 'Kurtosis', '`?e1071::kurtosis`',
  '`matrix`', 'Matrices', '`?base::matrix`',
  '`mean`', 'Arithmetic mean', '`?base::mean`',
  '`median`', 'Median', '`?stats::median`',
  '`Mode`', 'Mode', '`?pracma::Mode`',
  '`names`', 'Names of an object', '`?base::names`',
  '`nrow`', 'Number of rows', '`?base::nrow`',
  '`quantile`', 'Quantile', '`?stats::quantile`',
  '`sample`, `sample.int`', 'Random samples', '`?base::sample`; `?base::sample.int`', 
  '`sd`', 'Standard deviation', '`?stats::sd`',
  '`search`', 'Search path', '`?base::search`',
  '`seed`', 'Random number generation seed', '`?base::set.seed`',
  '`seq`, `seq.int`', 'Sequence generation', '`?base::seq`; `?base::seq.int`',
  '`skewness`', 'Skewness', '`?e1071::skewness`',
  '`sort`', 'Sorting vectors', '`?base::sort`',
  '`table`', 'Cross tabulation', '`?base::table`',
  '`var`', 'Variance', '`?stats::var`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable() 
```

# Chapter 1: *Introduction to Biostatistics*

## Example 1.4.1

Page 8 (10th ed), Page 7 (11th ed).

```{r}
EXA_C01_S04_01 |> 
  class() 
```

```{r}
EXA_C01_S04_01 |>
  dim()
```

```{r}
EXA_C01_S04_01 |> 
  head(n = 3L)
```

```{r}
EXA_C01_S04_01 |> 
  names()
```

```{r}
EXA_C01_S04_01$AGE
```

To answer Example 1.4.1,

```{r}
(nr = EXA_C01_S04_01 |> nrow())
set.seed(142); EXA_C01_S04_01[sample.int(n = nr, size = 10L), ]
```

## Example 1.4.2

Page 11 (10th ed), Page 10 (11th ed).

```{r}
EXA_C01_S04_01[seq.int(from = 4L, to = 166L, by = 18L), ]
```

# Chapter 2: *Descriptive Statistics*

## Example 2.2.1

Page 20 (10th ed), Page 19 (11th ed)

```{r}
EXA_C01_S04_01 |> 
  head(n = 3L)
```

```{r}
EXA_C01_S04_01$AGE |> 
  class()
```

Table 2.2.1.

```{r}
EXA_C01_S04_01$AGE |> 
  sort()
```

## Example 2.3.1

Page 23 (10th ed); Page 22 (11th ed).

Table 2.3.2

```{r}
EXA_C01_S04_01$AGE |>
  cut(breaks = seq.int(from=30, to=90, by=10), right = FALSE, ordered_result = TRUE) |>
  viewFreq()
```

## Example 2.4.1-2.4.6; 2.5.1-2.5.3

Page 38-42 (10th ed); Page 30-34 (11th ed).

Page 44-46 (10th ed); Page 35-41 (11th ed).

```{r}
EXA_C01_S04_01$AGE |> 
  viewStats()
```

## Example 2.5.4 (omitted)

Page 49 (10th ed), Page 41 (11th ed)

## Example 2.5.5

Page 50 (10th ed)

```{r}
EXA_C02_S05_05 |>
  head(n = 3L)
```

```{r}
EXA_C02_S05_05$GRF |>
  boxplot(main = c('Example 2.5.5 (10th ed)'))
```

```{r}
EXA_C02_S05_05$GRF |> 
  viewStats()
```

# Chapter 3: *Some Basic Probability Concepts*

## Example 3.4.1 - 3.4.9

Page 69-75 (10th ed), Page 61-67 (11th ed)

```{r}
c(28L, 19L, 41L, 53L, 35L, 38L, 44L, 60L) |>
  matrix(ncol = 2L, dimnames = list(
    FamilyHx = c('none', 'Bipolar', 'Unipolar', 'UniBipolar'), 
    Onset = c('Early', 'Late')
  )) |> 
  as.table() |>
  as_flextable()
```

## Example 3.5.1

Page 81 (10th ed), Page 72 (11th ed)

```{r}
d351 = c(495L, 14L, 5L, 436L) |>
  matrix(nrow = 2L, dimnames = list(
    Alzheimer = c('No', 'Yes'), 
    Test = c('Negative', 'Positive')
  )) |>
  binTab()
d351 |>
  as_flextable()
```

```{r}
d351 |>
  print(prevalence = .113, print_flextable = FALSE)
```

# Chapter 4: *Probability Distributions*

## Example 4.2.1-4.2.7

Page 93-97 (10th ed), Page 81-85 (11th ed)

```{r}
(d421 = rep(1:8, times = c(62L, 47L, 39L, 39L, 58L, 37L, 4L, 11L)))
```

Table 4.2.1, 4.2.2, Table 4.2.3

```{r}
d421 |>
  ordered() |>
  viewFreq()
```

## Example 4.2.8

Page 98 (10th ed), Page 85 (11th ed).

```{r}
viewStats(d421)
```

## Example 4.3.1

Page 99 (10th ed).

`d` for `density`; calculates $\text{Pr}(X = x)$.

```{r}
dbinom(x = 3L, size = 5L, prob = .858) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.3.1

Page 87 (11th ed).

```{r}
dbinom(x = 3L, size = 5L, prob = .899) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.3.2

Page 103 (10th ed), Page 90 (11th ed).

```{r}
dbinom(x = 4L, size = 10L, prob = .14) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.3.3

Page 103 (10th ed), Page 91 (11th ed).

`p` for `probability` or `distribution`; calculates

-   $\text{Pr}(X \leq x)$ if `lower.tail = TRUE` (default)
-   $\text{Pr}(X > x)$ if `lower.tail = FALSE`

```{r}
pL = pbinom(q = 5L, size = 25L, prob = .1, lower.tail = TRUE)
pL |> 
  label_percent(accuracy = .1)() |>
  noquote()
pU = pbinom(q = 5L, size = 25L, prob = .1, lower.tail = FALSE)
pU |> 
  label_percent(accuracy = .1)() |>
  noquote()
stopifnot(pL + pU == 1)
```

## Example 4.3.4

Page 105 (10th ed), Page 92 (11th ed)

```{r}
dbinom(x = 7L, size = 12L, prob = .55) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

```{r}
pbinom(q = 5L, size = 12L, prob = .55) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

```{r}
pbinom(q = 7L, size = 12L, prob = .55, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.4.1

Page 110 (10th ed), Page 97 (11th ed)

```{r}
dpois(x = 3L, lambda = 12) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.4.2

Page 110 (10th ed), Page 98 (11th ed)

```{r}
ppois(q = 2L, lambda = 12, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.4.3

Page 110 (10th ed), Page 98 (11th ed)

```{r}
ppois(q = 1L, lambda = 2) |> 
  label_percent(accuracy = .1)() |>
  noquote() 
```

## Example 4.4.4

Page 111 (10th ed), Page 98 (11th ed)

```{r}
dpois(x = 3L, lambda = 2) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.4.5

Page 112 (10th ed), Page 98 (11th ed)

```{r}
ppois(q = 5L, lambda = 2, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.6.1

Page 119 (10th ed), Page 106 (11th ed)

```{r}
pnorm(q = 2) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.6.2

Page 120 (10th ed), Page 106 (11th ed)

```{r}
(pnorm(2.55) - pnorm(-2.55)) |> 
  label_percent(accuracy = .1)() |>
  noquote()
(1 - 2 * pnorm(-2.55)) |> 
  label_percent(accuracy = .1)() |>
  noquote() # alternative solution
```

## Example 4.6.3

Page 121 (10th ed), Page 107 (11th ed)

```{r}
(pnorm(1.53) - pnorm(-2.74)) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.6.4

Page 121 (10th ed), Page 107 (11th ed)

```{r}
pnorm(2.71, lower.tail = FALSE) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.6.5

Page 122 (10th ed), Page 107 (11th ed)

```{r}
(pnorm(2.45) - pnorm(.84)) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.7.1

Page 122 (10th ed), Page 109 (11th ed)

```{r}
pnorm(q = 3, mean = 5.4, sd = 1.3) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.7.2

Page 125 (10th ed), Page 111 (11th ed)

```{r}
(pnorm(q = 649, mean = 491, sd = 119) - pnorm(q = 292, mean = 491, sd = 119)) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 4.7.3

Page 122 (10th ed), Page 111 (11th ed)

```{r}
pnorm(q = 8.5, mean = 5.4, sd = 1.3, lower.tail = FALSE) |> 
  label_percent(accuracy = .01)() |>
  noquote()
```

# Chapter 5, *Some Important Sampling Distributions*

## Example 5.3.2

Page 142 (10th ed)

```{r}
aggregated_z(xbar = 190, sd = 12.7, n = 10L, null.value = 185.6, alternative = 'greater')
```

## Example 5.3.3

Page 143 (10th ed)

```{r}
(pnorm(125, mean = 120, sd = 15/sqrt(50)) - pnorm(115, mean = 120, sd = 15/sqrt(50))) |> 
  label_percent(accuracy = .1)() |>
  noquote()
(aggregated_z(125, sd = 15, n = 50L, null.value = 120, alternative = 'less')$p.value - 
  aggregated_z(115, sd = 15, n = 50L, null.value = 120, alternative = 'less')$p.value) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

Page 127 (11th ed)

```{r}
(aggregated_z(110, sd = 20, n = 50L, null.value = 100, alternative = 'less')$p.value - 
  aggregated_z(95, sd = 20, n = 50L, null.value = 100, alternative = 'less')$p.value) |> 
  label_percent(accuracy = .1)() |>
  noquote()
```

## Example 5.4.1

Page 145 (10th ed), Page 129 (11th ed)

```{r}
aggregated_z(xbar = c(92, 105), sd = 20, n = 15L, null.value = 0, alternative = 'less') 
```

## Example 5.4.2

Page 148 (10th ed),

```{r}
aggregated_z(xbar = 20, sd = c(15, 20), n = c(35L, 40L), null.value = c(45, 30), 
  alternative = 'greater')
```

## Example 5.5.1

Page 150 (10th ed),

```{r}
prop_CLT(xbar = .4, n = 150L, null.value = .357, alternative = 'greater')
```

## Example 5.5.2

Page 152 (10th ed),

```{r}
prop_CLT(xbar = .45, n = 200L, null.value = .51, alternative = 'less')
```

## Example 5.6.1

Page 155 (10th ed),

```{r}
prop_CLT(xbar = .1, null.value = c(.28, .21), n = c(100L, 100L), alternative = 'greater')
```

## Example 5.6.2

Page 155 (10th ed),

```{r}
prop_CLT(xbar = .05, null.value = c(.34, .26), n = c(250L, 200L), alternative = 'less')
```

# Chapter 6, *Estimation*

## Example 6.2.1

Page 166 (10th ed), Page 147 (11th ed)

```{r}
aggregated_z(xbar = 22, n = 10L, sd = sqrt(45))
```

## Example 6.2.2

Page 168 (10th ed), Page 149 (11th ed)

```{r}
aggregated_z(xbar = 84.3, n = 15L, sd = sqrt(144), conf.level = .99)
```

## Example 6.2.3

Page 168 (10th ed), Page 150 (11th ed)

```{r}
aggregated_z(xbar = 17.2, n = 35L, sd = 8, conf.level = .9)
```

## Example 6.2.4

Page 169 (10th ed), Page 150 (11th ed)

```{r}
head(EXA_C06_S02_04)
aggregated_z(xbar = mean(EXA_C06_S02_04$ACTIVITY), n = nrow(EXA_C06_S02_04), sd = sqrt(.36))
```

## Example 6.3.1

Page 173 (10th ed),

```{r}
aggregated_t(xbar = 250.8, xsd = 130.9, n = 19L)
```

## Example 6.4.1

Page 177 (10th ed),

```{r}
aggregated_z(xbar = c(4.5, 3.4), sd = sqrt(c(1, 1.5)), n = c(12L, 15L))
```

## Example 6.4.2

Page 178 (10th ed),

```{r}
aggregated_z(xbar = c(4.3, 13), sd = c(5.22, 8.97), n = c(328L, 64L), conf.level = .99)
```

## Example 6.4.3

Page 180 (10th ed),

```{r}
aggregated_t(xbar = c(4.7, 8.8), xsd = c(9.3, 11.5), n = c(18L, 10L), var.equal = TRUE)
```

## Example 6.4.4

Page 181 (10th ed),

```{r}
aggregated_t(xbar = c(4.7, 8.8), xsd = c(9.3, 11.5), n = c(18L, 10L))
```

Welch slightly different from Cochran; textbook explained on Page 182

## Example 6.5.1

Page 185 (10th ed),

```{r}
prop_CLT(xbar = .18, n = 1220L)
```

## Example 6.6.1

Page 187 (10th ed),

```{r}
prop_CLT(x = c(31L, 53L), n = c(68L, 255L), conf.level = .99)
```

## Example 6.7.1

Page 190 (10th ed),

```{r}
uniroot(f = function(n, sd, level = .95) {
  qnorm(1-(1-level)/2) * sd/sqrt(n) - 5 # half-width of CI <= 5 grams
}, interval = c(0, 2e2), sd = 20)$root |>
  ceiling()
```

## Example 6.8.1

Page 192 (10th ed),

```{r}
uniroot(f = function(n, p, level = .95) {
  qnorm(1-(1-level)/2) * sqrt(p*(1-p)/n) - .05
}, interval = c(0, 1e3), p = .35)$root |>
  ceiling()
```

## Example 6.9.1

Page 196 (10th ed),

```{r}
d691 = c(9.7, 12.3, 11.2, 5.1, 24.8, 14.8, 17.7)
sqrt(aggregated_var(xsd = sd(d691), n = length(d691)))
```

## Example 6.10.1

Page 200 (10th ed),

```{r}
aggregated_var(xsd = c(8.1, 5.9), n = c(16L, 4L))
```

# Chapter 7: *Hypothesis Testing*

## Example 7.2.1

Page 222 (10th ed); Page 201 (11th ed)

```{r}
aggregated_z(xbar = 27, sd = sqrt(20), n = 10L, null.value = 30)
```

## Example 7.2.2

Page 226 (10th ed); Page 204 (11th ed)

```{r}
aggregated_z(xbar = 27, sd = sqrt(20), n = 10L, null.value = 30, alternative = 'less')
```

## Example 7.2.3

Page 228 (10th ed); Page 206 (11th ed)

```{r}
EXA_C07_S02_03 |> head(n = 3L)
t.test(EXA_C07_S02_03$DAYS, mu = 15)
```

## Example 7.2.4

Page 231 (10th ed); Page 209 (11th ed)

```{r}
aggregated_z(xbar = 146, sd = 27, n = 157L, null.value = 140, alternative = 'greater')
```

## Example 7.2.5

Page 232 (10th ed); Page 210 (11th ed)

```{r}
d725 = c(33.38, 32.15, 34.34, 33.95, 33.46, 34.13, 33.99, 34.10, 33.85, 
  34.23, 34.45, 34.19, 33.97, 32.73, 34.05)
t.test(d725, mu = 34.5)
```

## Example 7.3.1

Page 237 (10th ed), Page 213 (11th ed)

```{r}
aggregated_z(xbar = c(4.5, 3.4), sd = sqrt(c(1, 1.5)), n = c(12L, 15L), null.value = 0)
```

## Example 7.3.2

Page 239 (10th ed), Page 215 (11th ed)

```{r}
EXA_C07_S03_02 |> 
  head(n = 3L)
```

```{r}
EXA_C07_S03_02 |>
  with(t.test(x = CONTROL, y = SCI, alternative = 'less', var.equal = TRUE))
```

## Example 7.3.3

Page 240 (10th ed), Page 217 (11th ed)

```{r}
aggregated_t(xbar = c(19.16, 9.53), xsd = c(5.29, 2.69), n = c(15L, 30L), null.value = 0)
```

## Example 7.3.4

Page 242 (10th ed), Page 219 (11th ed)

```{r}
aggregated_z(xbar = c(59.01, 46.61), sd = c(44.89, 34.85), n = c(53L, 54L), null.value = 0,
  alternative = 'greater')
```

## Example 7.4.1

Page 251 (10th ed), Page 226 (11th ed)

```{r}
EXA_C07_S04_01 |> head(n = 3L)
```

```{r}
EXA_C07_S04_01 |>
  with(expr = t.test(x = POSTOP, y = PREOP, alternative = 'greater', paired = TRUE))
```

## Example 7.5.1

Page 258 (10th ed), Page 232 (11th ed)

```{r}
prop_CLT(x = 24L, n = 301L, null.value = .063, alternative = 'greater') 
```

## Example 7.6.1

Page 261 (10th ed), Page 235 (11th ed)

```{r}
prop_CLT(x = c(24L, 11L), n = c(44L, 29L), null.value = 0, alternative = 'greater')
```

## Example 7.7.1

Page 264 (10th ed), Page 238 (11th ed)

```{r}
EXA_C07_S07_01 |> 
  head(n = 3L)
```

```{r}
aggregated_var(xsd = sd(EXA_C07_S07_01$mass), n = 16L, null.value = 600)
```

## Example 7.8.1

Page 268 (10th ed), Page 242 (11th ed)

```{r}
aggregated_var(xsd = c(30.62, 11.37), n = 6L, null.value = 1, alternative = 'greater')
```

## Example 7.8.2

Page 270 (10th ed),

```{r}
EXA_C07_S03_02 |>
  with(var.test(x = CONTROL, y = SCI))
```

## Example 7.9.1

Page 272 (10th ed), Page 245 (11th ed). Table 7.9.1

```{r}
seq.int(from = 16, to = 19, by = .5) |>
  power_z(null.value = 17.5, sd = 3.6, n = 100L)
```

## Example 7.9.2

Page 276 (10th ed), Page 248 (11th ed)

```{r}
seq.int(from = 50, to = 70, by = 5) |>
   power_z(null.value = 65, sd = 15, n = 20L, sig.level = .01, alternative = 'less')
```

## Example 7.10.1

Page 278 (10th ed),

```{r}
(n_d7101 <- uniroot(f = function(x) {
  power_z(55, null.value = 65, sd = 15, n = x, sig.level = .01, alternative = 'less')$power - .95
}, interval = c(0, 50))$root)
power_z(55, null.value = 65, sd = 15, n = ceiling(n_d7101), sig.level = .01, alternative = 'less')
```

# Chapter 8: *Analysis of Variance*

## Example 8.2.1

Page 318 (10th ed), Page 280 (11th ed)

```{r}
EXA_C08_S02_01 |> head(n = 3L)
```

```{r}
boxplot(selenium ~ type, data = EXA_C08_S02_01, main = 'Figure 8.2.7')
(aov_d821 = aov(selenium ~ type, data = EXA_C08_S02_01)) 
# ?stats::aov # analysis-of-variance model
anova(aov_d821) 
# ?stats::anova # ANOVA table
```

## Example 8.2.2

Page 325 (10th ed), Page 286 (11th ed)

```{r}
(tukey_d822 <- TukeyHSD(aov_d821, conf.level = 0.95)) # Figure 8.2.8
plot(tukey_d822)
```

## Example 8.3.1

Page 339 (10th ed), Page 298 (11th ed)

```{r}
head(EXA_C08_S03_01)
head(d831 <- within(EXA_C08_S03_01, expr = {
  ageGroup = structure(ageGroup, levels = c('<20', '20s', '30s', '40s', '>50'), class = 'factor')
}))
(aov_831 = aov(time ~ method + ageGroup, data = d831))
anova(aov_831)
```

## Example 8.4.1

Page 348 (10th ed), Page 307 (11th ed)

```{r}
head(EXA_C08_S04_01)
head(d841 <- within(EXA_C08_S04_01, expr = {
  SUBJ = factor(SUBJ)
  TIME = structure(TIME, levels = c('Baseline', '1-Mon', '3-Mon', '6-Mon'), class = 'factor')
}))
(aov_841 = aov(FUNC ~ SUBJ + TIME, data = d841))
anova(aov_841)
```

## Example 8.4.2

Page 352 (10th ed), Page 310 (11th ed). Optional; out of the scope of this course.

```{r}
head(EXA_C08_S04_02)
names(EXA_C08_S04_02)[3:6] = c('baseline', '2wk', '4wk', '6wk')
head(d842a <- within(EXA_C08_S04_02, expr = {
  subject = factor(subject)
  treatment = structure(treatment, levels = c('placebo', 'aloe_juice'), class = 'factor')
}))
head(d842b <- reshape2::melt(d842a, id.vars = c('subject', 'treatment'), 
                             variable.name = 'time', value.name = 'OralScores'))
# Hypothesis: 
# Main effect of 'treatment';
# Main effect of 'time';
# Interaction between 'treatment' and 'time'
(aov_842 = aov(OralScores ~ treatment * time + Error(subject), data = d842b))
class(aov_842)
summary(aov_842)
# Section 'Error: subject' in R output
# .. is Figure 8.4.4 'Tests of Between-Subjects Effects' (without the row of 'Intercept')
# .. 'treatment' row: effect of treatment at **baseline** (i.e. reference time), 
# ... degree-of-freedom (dof) = 2-1 = 1
# .. 'Residuals' row: residual at baseline, dof = (25-1) - (2-1) = 23
# .. It's important to note that 'treatment' is a **between-subject factor**.
# Section 'Error: Within' in R output
# .. is Figure 8.4.4 'Tests of Within-Subjects Effects'
# .. 'time' row: effect of time within subject for placebo (i.e. reference treatment), dof = 4-1 = 3
# .. 'treatment:time' row: interation of treatment and time, dof = (2-1)*(4-1) = 3
# .. 'Residuals' row: residual at 2wk, 4wk and 6wk, dof = (4-1)*23 = 69 
# ... [(4-1) timepoints, 23 dof at each timepoints]
# Analysis Interpretation
# .. No signif. diff. detected between placebo vs. aloe at baseline (p = .815)
# .. No signif. diff. detected in the trends over time between placebo vs. aloe (p = .974)
# .. Signif. diff. detected among the four timepoints, for either placebo or aloe pts (p = 3e-7)
# R code below creates an equivalent ANOVA model
anova(aov(OralScores ~ treatment * time + subject, data = d842b))
# .. 'subject' is considered as a block factor 
```

## Example 8.5.2

Page 364 (10th ed), Page 321 (11th ed)

```{r}
head(EXA_C08_S05_02)
head(d852 <- within(EXA_C08_S05_02, expr = {
  A = structure(A, levels = c('Cardiac', 'Cancer', 'CVA', 'Tuberculosis'), class = 'factor')
  B = structure(B, levels = c('20s', '30s', '40s', '50+'), class = 'factor')
}))
(aov_852 = aov(HOME ~ A * B, data = d852))
anova(aov_852)
summary(lm(HOME ~ A * B, data = d852)) 
# produces alpha, beta and (alpha beta)'s in the formulation 
```

# Chapter 9: *Simple Linear Regression and Correlation*

## Example 9.3.1

Page 417 (10th ed), Page 358 (11th ed)

```{r}
EXA_C09_S03_01 |> head(n = 3L)
names(EXA_C09_S03_01)[2:3] = c('Waist', 'AT')
```

```{r}
m931 = lm(AT ~ Waist, data = EXA_C09_S03_01)
plot(AT ~ Waist, data = EXA_C09_S03_01, 
     xlab = 'Waist circumference (cm), X', 
     ylab = 'Deep abdominal AT area (cm2), Y', 
     main = 'Figure 9.3.1 and 9.3.3'); abline(m931, col = 'red', lty = 2)
```

## Example 9.4.1

Page 432 (10th ed), Page 372 (11th ed) Table 9.4.1. Page 434 (10th ed).

```{r}
anova(m931)
```

## Example 9.4.2

Page 436 (10th ed), Page 375 (11th ed)

```{r}
summary(m931)
```

Page 438 (10th ed).

```{r}
confint(m931)
```

## Example 9.4.3

Page 376 (11th ed). Omitted

## Example 9.4.3; 9.4.4

Page 440 (10th ed). Page 379 (11th ed)

```{r}
plot(m931, which = 1, main = 'Figure 9.4.8 (10th) or 9.4.9 (11th)')
```

## Example 9.7.1

Page 447 (10th ed), Page 386 (11th ed)

```{r}
EXA_C09_S07_01 |> head(n = 3L)
```

```{r}
plot(CV ~ HEIGHT, data = EXA_C09_S07_01, 
     xlab = 'Height (cm)', 
     ylab = 'Cv (units)', 
     main = 'Figure 9.7.2')
m971 = lm(CV ~ HEIGHT, data = EXA_C09_S07_01)
summary(m971)
```

Figure 9.7.3; Page 450 (10th ed)

```{r}
anova(m971)
```

## Example 9.7.2

Page 452 (10th ed), Page 390 (11th ed)

```{r}
cor(EXA_C09_S07_01); 
```

Figure 9.7.4, 9.7.5.

```{r}
cor.test(~ CV + HEIGHT, data = EXA_C09_S07_01)
```

Page 453, when the hypothesized rho ss a nonzero value; R does not have a function to do this.

# Chapter 10: *Multiple Regression and Correlation*

## Example 10.3.1; 10.4.1-10.4.3

Page 493 (10th ed), Page 419 (11th ed)

Page 502 (10th ed), Page 428 (11th ed)

Page 504 (10th ed), Page 429 (11th ed)

Page 505 (10th ed), Page 430 (11th ed)

Page 506 (10th ed), Page 431 (11th ed)

```{r}
EXA_C10_S03_01 |> head(n = 3L)
```

```{r}
pairs(EXA_C10_S03_01, main = 'Figure 10.3.1')
```

see 'Multiple R-squared' (not 'Adjusted R-squared')

see 'F-statistic'

see 'Coefficients:'

```{r}
m1031 = lm(CDA ~ AGE + EDLEVEL, data = EXA_C10_S03_01)
summary(m1031)
```

confidence interval for beta's;

```{r}
confint(m1031)
```

## Example 10.5.1

Page 509 (10th ed), Page 434 (11th ed)

```{r}
(newd_1031 = data.frame(AGE = 68, EDLEVEL = 12))
predict(m1031, newdata = newd_1031, interval = 'prediction')
predict(m1031, newdata = newd_1031, interval = 'confidence')
```

## Example 10.6.1

Page 511 (10th ed), Page 436 (11th ed)

```{r}
head(EXA_C10_S06_01)
pairs(EXA_C10_S06_01, main = 'Scatter Plot Matrix, Example 10.6.1')
m1061 = lm(W ~ P + S, data = EXA_C10_S06_01)
summary(m1061)
confint(m1061)
anova(m1061)
```

## Example 10.6.2;

Page 515 (10th ed), Page 440 (11th ed)

```{r}
psych::partial.r(EXA_C10_S06_01, x = 2:3, y = 1L)
```

# Chapter 11: *Regression Analysis: Some Additional Techniques*

## Example 11.1.1

Page 540,

```{r}
head(EXA_C11_S01_01)
head(log(EXA_C11_S01_01$conc, base = 10))
head(EXA_C11_S01_01$logConc)
```

## Example 11.1.2

Page 542,

```{r}
head(EXA_C11_S01_02)
```

```{r}
cor.test(~ sbp + weight, data = EXA_C11_S01_02)
```

```{r}
cor.test(~ sbp + bmi, data = EXA_C11_S01_02) 
```

## Example 11.2.1

Page 545,

```{r}
head(EXA_C11_S02_01)
d1121 = EXA_C11_S02_01 |>
  within(expr = {
    SMOKE = as.logical(SMOKE)
  })
xlab1121 = 'Length of gestation (weeks)'; ylab1121 = 'Birth weight (grams)'
car::scatterplot(GRAMS ~ WEEKS | SMOKE, data = d1121, regLine = FALSE, smooth = FALSE,
            xlab = xlab1121, ylab = ylab1121, main = 'Figure 11.2.1')
m1121 = lm(GRAMS ~ WEEKS + SMOKE, data = d1121)
summary(m1121) # Figure 11.2.2
confint(m1121)
car::scatterplot(GRAMS ~ WEEKS | SMOKE, data = d1121, regLine = FALSE, smooth = FALSE,
            xlab = xlab1121, ylab = ylab1121, main = 'Figure 11.2.3')
(cf1121 = m1121$coefficients)
abline(a = cf1121[1L], b = cf1121[2L], col = 'blue') # regression line for non-smoking mothers
abline(a = cf1121[1L] + cf1121[3L], b = cf1121[2L], col = 'magenta') 
```

## Example 11.2.3; Page 551,

```{r}
d1123 = within(EXA_C11_S02_03, expr = {
  METHOD = factor(METHOD, levels = c('C', 'A', 'B')) # textbook designated 'C' as reference level
})
summary(m1123 <- lm(EFFECT ~ AGE * METHOD, data = d1123)) # Figure 11.2.5
confint(m1123)
car::scatterplot(EFFECT ~ AGE | METHOD, data = d1123, smooth = FALSE,
            xlab = 'Age', ylab = 'Treatment effectiveness', main = 'Figure 11.2.6')
```

## Example 11.3.1

Page 561,

```{r}
head(EXA_C11_S03_01)
names(EXA_C11_S03_01) = c('JOBPER', 'ASRV', 'ENTH', 'AMB', 'COMM', 'PROB', 'INIT')
summary(m1131_raw <- lm(JOBPER ~ ASRV + ENTH + AMB + COMM + PROB + INIT, data = EXA_C11_S03_01))
# summary(m1131 <- MASS::stepAIC(m1131_raw, direction = 'backward'))
# the stepwise selection criterion used in MINITAB is not necessarily AIC
```

## Example 11.4.1

Page 572,

```{r}
d1141 = as.table(array(c(92L, 21L, 15L, 20L), dim = c(2L, 2L), dimnames = list(
  OCAD = c('Present', 'Absent'), Sex = c('Male', 'Female'))))
as_flextable(d1141) # Table 11.4.2
(d1141a = within(as.data.frame.table(as.table(d1141)), expr = {
  OCAD = (OCAD == 'Present')
  Sex = factor(Sex, levels = c('Female', 'Male'))
}))
m1141 = glm(OCAD ~ Sex, family = binomial, weights = Freq, data = d1141a)
summary(m1141) # Figure 11.4.1
exp(m1141$coefficients[2L]) # exp(beta_M)
exp(confint(m1141)) # confidence interval of exp(beta)
predict(m1141, newdata = data.frame(Sex = setNames(nm = c('Male', 'Female'))), type = 'response')
```

## Example 11.4.2

Page 573,

```{r}
head(EXA_C11_S04_02)
summary(m1142 <- glm(ATT ~ AGE, family = binomial, data = EXA_C11_S04_02)) # Figure 11.4.2
exp(m1142$coefficients[2L])
exp(confint(m1142))
car::Anova(m1142) # Optional
```

## Example 11.4.3

Page 576,

```{r}
head(REV_C11_24)
summary(glm(ONSET ~ HIAA + TRYPT, family = binomial, data = REV_C11_24)) # Figure 11.4.4
# Predictor TRYPT should be removed from model due to p-value \approx 1 
summary(glm(ONSET ~ HIAA, family = binomial, data = REV_C11_24)) 
```

## Example 11.4.4-11.4.5

Page 578-579

```{r}
DescTools::PseudoR2(m1142, which = 'CoxSnell')
DescTools::PseudoR2(m1142, which = 'Nagelkerke')
```

# Chapter 12:

## Example 12.3.1

Page 605 (10th ed)

```{r}
d1231_b = c(-Inf, seq.int(from = 125, to = 275, by = 25), Inf)
d1231 = c(1L, 3L, 8L, 18L, 6L, 4L, 4L, 3L) |>
  setNames(nm = cut(double(), breaks = d1231_b, right = FALSE, include.lowest = TRUE) |> levels())
chi1231 = viewOE(d1231, prob = diff.default(pnorm(q = d1231_b, mean = 198.67, sd = 41.31)))
pchisq(sum(chi1231), df = length(d1231) - 3L, lower.tail = FALSE)
```

-3L: three restrictions (explained on Page 608) (1) making sum(xo) == sum(xe) (2) estimating mean (3) estimating sd

## Example 12.3.2

Page 609 (10th ed),

```{r}
# 100 doctors, 25 patients per doctor
d1232 = c(5L, 6L, 8L, 10L, 10L, 15L, 17L, 10L, 10L, 9L, 0L)
o1232 = setNames(c(sum(d1232[1:2]), d1232[-(1:2)]), nm = c('0-1', 2:9, '10 or more'))
(p1232 = sum((0:10) * d1232) / (25 * 100)) # binomial `prob`
chi1232 = viewOE(o1232, prob = c(
  pbinom(1L, size = 25L, prob = p1232),
  dbinom(2:9, size = 25L, prob = p1232),
  pbinom(9, size = 25L, prob = p1232, lower.tail = FALSE)))
pchisq(sum(chi1232), df = length(o1232) - 2L, lower.tail = FALSE)
```

-2L: two restrictions (explained on Page 611) (1) making sum(o) == sum(e) (2) estimating p1232

## Example 12.3.3

Page 611 (10th ed),

```{r}
d1233 = c(5L, 14L, 15L, 23L, 16L, 9L, 3L, 3L, 1L, 1L, 0L)
o_1233 = setNames(c(d1233[1:8], sum(d1233[-(1:8)])), nm = c(0:7, '8 or more'))
p_1233 = c(dpois(0:7, lambda = 3), # lambda = 3 is provided by the textbook
           ppois(7L, lambda = 3, lower.tail = FALSE))
chi1233 = viewOE(o_1233, prob = p_1233)
pchisq(sum(chi1233), df = length(o_1233) - 1L, lower.tail = FALSE)
```

-1L: one restrictions (1) making sum(xo) == sum(xe)

```{r}
chisq.test(o_1233, p = p_1233) # equivalent # warning on any(E < 5)
```

## Example 12.3.4

Page 614 (10th ed), Page 531 (11th ed)

```{r}
d1234 = c('Dec 05' = 62L, 'Jan 06' = 84L, 'Feb 06' = 17L, 'Mar 06' = 16L, 'Apr 06' = 21L)
viewOE(d1234) # Figure 12.3.2
chisq.test(d1234) 
```

## Example 12.3.5

Page 616 (10th ed), Page 533 (11th ed)

```{r}
d1235 = c(dominant = 43L, heterozygous = 125L, recessive = 32L)
viewOE(d1235, prob = c(1, 2, 1))
chisq.test(d1235, p = c(1, 2, 1), rescale.p = TRUE)
```

## Example 12.4.1

Page 621 (10th ed)

```{r}
d1241 = as.table(array(c(260L, 15L, 7L, 299L, 41L, 14L), dim = c(3L, 2L), dimnames = list(
  Race = c('White', 'Black', 'Other'), FolicAcid = c('TRUE', 'FALSE'))))
as_flextable(d1241)
chisq.test(d1241) # ?stats::chisq.test
```

## Example 12.4.2

Page 626 (10th ed),

```{r}
d1242 = as.table(array(c(131L, 14L, 52L, 36L), dim = c(2L, 2L), dimnames = list(
  Type = c('Faller', 'Non-Faller'), LifestyleChange = c('TRUE', 'FALSE'))))
as_flextable(d1242)
chisq.test(d1242, correct = FALSE)
chisq.test(d1242, correct = TRUE) # Yates's Correction
```

## Example 12.5.1

Page 631 (10th ed),

```{r}
d1251 = as.table(array(c(21L, 19L, 75L, 77L), dim = c(2L, 2L), dimnames = list(
  Group = c('Narcoleptic', 'Healthy'), Migraine = c('TRUE', 'FALSE'))))
as_flextable(d1251)
(chisq_1251 = chisq.test(d1251, correct = FALSE))
if (FALSE) {
  # (optional) using test on two proportions
  # only equivalent for 2*2 contingency table
  (clt_1251 = prop_CLT(x = c(21L, 19L), n = 96L, null.value = 0))
  all.equal.numeric(unname(clt_1251$statistic^2), unname(chisq_1251$statistic))
}
```

## Example 12.6.1

Page 638 (10th ed),

```{r}
d1262 = as.table(array(c(2L, 8L, 7L, 4L), dim = c(2L, 2L), dimnames = list(
  Group = c('PI Naive', 'PA Experienced'), Regimen2yr = c('TRUE', 'FALSE'))))
as_flextable(d1262)
fisher.test(d1262)
```

## Example 12.7.1

Page 644 (10th ed),

```{r}
d1271 = as.table(array(c(22L, 18L, 216L, 199L), dim = c(2L, 2L), dimnames = list(
  Exercising = c('Extreme', 'No'), PretermLabor = c('Cases', 'Noncases'))))
as_flextable(d1271)
DescTools::RelRisk(d1271, conf.level = .95) # 95% CI for relative risk
fisher.test(d1271) # 95% CI for odds ratio
```

## Example 12.7.2

Page 647 (10th ed),

```{r}
d1272 = as.table(array(c(64L, 68L, 342L, 3496L), dim = c(2L, 2L), dimnames = list(
  SmkPregnancy = c('Smoked Throughout', 'Never Smoked'), Obesity = c('Cases', 'Noncases'))))
as_flextable(d1272)
fisher.test(d1272)
```

## Example 12.7.3-12.7.4

Page 650-652 (10th ed),

```{r}
d1273 = array(c(21L, 16L, 11L, 6L, 50L, 18L, 14L, 6L), dim = c(2L, 2L, 2L), dimnames = list(
  HTN = c('Present', 'Absent'), OCAD = c('Cases', 'Controls'), Age = c('<=55', '>55')))
addmargins(d1273, margin = 1:2) # Table 12.7.6
mantelhaen.test(d1273)
```

# Chapter 13: *Nonparametric and Distribution-Free Statistics*

## Example 13.3.1

Page 673

```{r}
d1331 = c(4, 5, 8, 8, 9, 6, 10, 7, 6, 6)
BSDA::SIGN.test(d1331, md = 5)
```

## Example 13.3.2

Page 677,

```{r}
head(EXA_C13_S03_02, n = 3L)
EXA_C13_S03_02 |> with(expr = {
  BSDA::SIGN.test(x = X, y = Y, alternative = 'less')
})
```

## Example 13.4.1

Page 683

```{r}
d1341 = c(4.91, 4.10, 6.74, 7.27, 7.42, 7.50, 6.56, 4.64, 5.98, 3.14, 3.23, 5.80, 6.17, 5.39, 5.77)
wilcox.test(d1341, mu = 5.05)
```

## Example 13.5.1

Page 686,

```{r}
head(EXA_C13_S05_01, n = 3L)
(med1351 = EXA_C13_S05_01 |> unlist() |> median(na.rm = TRUE)) # common median
t1351 = EXA_C13_S05_01 |> with(expr = {
  tmp = cbind(Urban = table(URBAN < med1351), Rural = table(RURAL < med1351, useNA = 'no'))
  rownames(tmp) = paste('Number of scores', c('above', 'below'), 'median')
  as.table(tmp)
}) 
as_flextable(t1351) # Table 13.5.2
chisq.test(t1351, correct = FALSE)
```

## Example 13.6.1;

Page 691,

```{r}
head(EXA_C13_S06_01)
with(EXA_C13_S06_01, wilcox.test(X, Y, exact = FALSE, alternative = 'less'))
```

## Example 13.7.1;

Page 699,

```{r}
head(EXA_C13_S07_01)
ks.test(EXA_C13_S07_01$GLUCOSE, y = pnorm, mean = 80, sd = 6)
```

## Example 13.8.1;

Page 705 (10th ed), Page 611 (11th ed)

```{r}
(d1381 = data.frame(Air = c(12.22, 28.44, 28.13, 38.69, 54.91),
                    Benzaldehyde = c(3.68, 4.05, 6.47, 21.12, 3.33),
                    Acetaldehyde = c(54.36, 27.87, 66.81, 46.27, 30.19))) # Table 13.8.1
with(data = reshape2::melt(
  data = d1381, 
  id.vars = NULL, 
  value.name = 'Eosinophil', 
  variable.name = 'Exposure'
), expr = kruskal.test(x = Eosinophil, g = Exposure))
```

## Example 13.8.2;

Page 708 (10th ed), Page 613 (11th ed)

```{r}
head(EXA_C13_S08_02)
with(data = reshape2::melt(
  data = EXA_C13_S08_02, 
  id.vars = NULL, 
  value.name = 'Value', 
  variable.name = 'Hospital'
), expr = kruskal.test(x = Value, g = Hospital))
```

## Example 13.9.1;

Page 713 (10th ed); Page 618 (11th ed)

```{r}
head(EXA_C13_S09_01)
m1391 = as.matrix(EXA_C13_S09_01[LETTERS[1:3]], rownames.force = TRUE)
names(dimnames(m1391)) = c('Therapist', 'Model')
m1391 # Table 13.9.1
friedman.test(m1391)
```

## Example 13.9.2;

Page 715 (10th ed); Page 620 (11th ed)

```{r}
head(EXA_C13_S09_02)
m1392 = as.matrix(EXA_C13_S09_02[LETTERS[1:4]], rownames.force = TRUE)
names(dimnames(m1392)) = c('Animal', 'Dose')
m1392 # Table 13.9.2
friedman.test(m1392)
```

## Example 13.10.1;

Page 720,

```{r}
head(EXA_C13_S10_01, n = 3L)
with(EXA_C13_S10_01, cor.test(X, Y, method = 'spearman'))
```

## Example 13.10.2;

Page 722,

```{r}
head(EXA_C13_S10_02, n = 3L)
with(EXA_C13_S10_02, cor.test(V2, V3, method = 'spearman', exact = FALSE))
```

## Example 13.11.1-13.11.2;

Page 728-729,

```{r}
testosterone <- c(230, 175, 315, 290, 275, 150, 360, 425)
citricAcid <- c(421, 278, 618, 482, 465, 105, 550, 750)
robslopes::TheilSen(x = citricAcid, y = testosterone)
# textbook uses *central* median of ordered pairs, while ?robslopes::TheilSen uses *upper* median
mblm::mblm(testosterone ~ citricAcid, repeated = FALSE) |> summary() # Figure 13.11.1 (11th ed)
```

# Chapter 14: *Survival Analysis*

## Example 14.3.1

Page 756 (10th ed), Page 652 (11th ed).

```{r}
EXA_C14_S03_01 |> head(n = 3L)
d1431 = EXA_C14_S03_01 |> 
  within(expr = {
    TIME = .difftime(TIME, units = 'months')
    OS = Surv(time = TIME, event = (VITAL != 'ned'))
  })
class(d1431$OS)
head(d1431$OS)
```

Table 14.3.2

```{r}
summary(sf_1431 <- survfit(OS ~ TUMOR, data = d1431))
```

Figure 14.3.1

```{r}
sf_1431 |> 
  plot(lty = 2:3, xlab = 'month', ylab = 'OS', main = 'Figure 14.3.1')
legend(x = 250, y = .9, legend = c('High', 'Low'), lty = 2:3)
```

Fancy plot via `?ggfortify:::autoplot.survfit`

```{r}
#| warning: false
sf_1431 |> 
  autoplot() + xlim(0, 120) + labs(title = 'Figure 14.3.1')
```

Fancy plot via `?survminer:::ggsurvplot`

```{r}
sf_1431 |> 
  ggsurvplot(title = 'Figure 14.3.1')
```

## Example 14.4.1

Page 764 (10th ed), Page 659 (11th ed)

```{r}
survdiff(OS ~ TUMOR, data = d1431)
```

## Example 14.5.1

Page 769 (10th ed), Page 663 (11th ed)

```{r}
head(EXA_C14_S05_01)
head(d1451 <- within(EXA_C14_S05_01, expr = {
  time = .difftime(time, units = 'weeks')
  PFS = Surv(time, status)
  drug = relevel(structure(drug, levels = c('Opiate', 'Other'), class = 'factor'), ref = 'Other')
}))
```

```{r}
summary(model1_1451 <- coxph(PFS ~ drug + age, data = d1451))
```

'Opiate' has higher hazard compared to Drug='Other' (hazard ratio (HR) = 9.407, p \< .001) With 'proportional hazard' assumption, we don't need to discuss Opiate vs. Other at any specific time point

```{r}
confint(model1_1451)
```

'age' is not significant (p = .772), so it should be removed from the model

```{r}
summary(model2_1451 <- coxph(PFS ~ drug, data = d1451))
confint(model2_1451)
```

'Opiate' has higher hazard compared to Drug='Other' (hazard ratio (HR) = 9.923, p \< .001)

plot

```{r}
sf_1451 = survfit(PFS ~ drug, data = d1451)
sf_1451 |> plot(lty = 2:3, xlab = 'weeks', ylab = 'PFS', main = 'Figure 14.5.1')
legend(x = 35, y = .9, legend = c('Other', 'Opiate'), lty = 2:3)
```

fancy plot

```{r}
sf_1451 |> autoplot() + labs(title = 'Figure 14.5.1')
```

another fancy plot

```{r}
sf_1451 |> ggsurvplot(title = 'Figure 14.5.1')
```
